---
title: "Table S9 — Correlation of CMAP activity with gestational and post-menstrual age within clinical subgroups"
subtitle: "Spearman correlations of subject-level CoGAPS CMAP activity with GA and PMA, by clinical stratum"
author: "Kratimenos Lab"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
suppressMessages({library(readxl); library(dplyr); library(knitr); library(kableExtra)})

find_data_dir <- function(sub = ".", max_up = 6) {
  prefix <- "."
  for (i in 0:max_up) { cand <- file.path(prefix, "Supplementary_Datasets", sub); if (dir.exists(cand)) return(cand)
    prefix <- file.path(prefix, "..") }
  stop("Supplementary_Datasets/", sub, " not found above ", normalizePath("."))
}
DIR  <- find_data_dir()
WB   <- file.path(DIR, "Data_S3_SpatialTx_Tables.xlsx")   # consolidated: correlations + metrics + subgroups
mn   <- function(x) gsub("-", "−", sprintf("%.3f", x))   # 3 dp, typographic minus
```

# Overview

Subject-level CoGAPS cerebellar maturation-associated pattern (CMAP) activity is correlated (Spearman)
with **gestational age at birth (GA)** and **post-menstrual age at death (PMA)** across the spatial
transcriptomic cohort and within clinical strata. Patterns map as **H-16 = CMAP_IGL_IMM**,
**H-17 = CMAP_IGL**, **H-27 = CMAP_PCL**. Coefficients are descriptive (n small); no P values are reported.

Subgroup analyses are restricted to clinical strata containing **≥ 5 subjects**, since correlation
estimates become unstable at smaller sample sizes; strata below that threshold are not shown.

# Data

All inputs are consolidated in `Data_S3_SpatialTx_Tables.xlsx` (one sheet each):
* sheets `ga_correlations`, `pma_correlations` — Spearman ρ per subgroup for each pattern (H-16/17/27),
  from the CoGAPS subgroup analysis (subject-level CMAP activity vs GA / vs PMA).
* sheet `spatial_cohort_subgroups` — de-identified per-subject subgroup membership, used only to count each
  stratum's *n*.
* sheet `allsubjects_cmap_metrics` — de-identified per-subject CMAP activity + GA + PMA for the 11-subject
  full cohort (from the cross-modal concordance analysis; 2099 excluded). The **All-subjects (full-cohort)
  row** is computed from it here (Spearman ρ of each pattern vs GA and vs PMA) — not taken from the
  subgroup-restricted correlation files.

```{r load}
rowlab <- function(sheet) { d <- as.data.frame(read_excel(WB, sheet = sheet)); rownames(d) <- d[[1]]; d[[1]] <- NULL; d }
ga  <- rowlab("ga_correlations")    # subgroup-row x pattern-col Spearman matrix (first col = row label)
pma <- rowlab("pma_correlations")
meta <- read_excel(WB, sheet = "spatial_cohort_subgroups")
asm  <- as.data.frame(read_excel(WB, sheet = "allsubjects_cmap_metrics"))   # per-subject CMAP + GA + PMA (n = 11)

# subgroup code (row in the correlation files) -> display label + membership rule (for n)
SUBGROUPS <- tibble::tribble(
  ~code,                          ~label,                                    ~rule,
  "preterm 1",                    "Preterm",                                 quote(preterm == 1),
  "sex_male 1",                   "Male",                                    quote(sex_male == 1),
  "sex_male 0",                   "Female",                                  quote(sex_male == 0),
  "nec 0",                        "Without NEC",                             quote(nec == 0),
  "sepsis 0",                     "Without sepsis",                          quote(sepsis == 0),
  "chorio_or_infection_any 0",    "Without chorioamnionitis or infection",   quote(chorio_or_infection_any == 0),
  "preeclampsia 0",               "Without preeclampsia",                    quote(preeclampsia == 0),
  "hypoxia_vent 0",               "Without ventilatory support",             quote(hypoxia_vent == 0),
  "spontaneous_birth 1",          "With spontaneous onset of labor",         quote(spontaneous_birth == 1))

# full-cohort ('All subjects') row — computed here: Spearman ρ of each pattern vs GA and vs PMA (n = 11)
scor <- function(pat, age) round(unname(suppressWarnings(cor(asm[[pat]], asm[[age]], method = "spearman"))), 3)
ALL_SUBJECTS <- list(
  label = "All subjects", n = nrow(asm),
  IGL_IMM_GA = scor("CMAP_IGL_IMM","GA_wk"), IGL_IMM_PMA = scor("CMAP_IGL_IMM","PMA_wk"),
  IGL_GA     = scor("CMAP_IGL","GA_wk"),     IGL_PMA     = scor("CMAP_IGL","PMA_wk"),
  PCL_GA     = scor("CMAP_PCL","GA_wk"),     PCL_PMA     = scor("CMAP_PCL","PMA_wk"))
```

# Table S9

```{r tableS9, echo=FALSE}
rows <- list(with(ALL_SUBJECTS, c(label, as.character(n),
             mn(IGL_IMM_GA), mn(IGL_IMM_PMA), mn(IGL_GA), mn(IGL_PMA), mn(PCL_GA), mn(PCL_PMA))))
for (i in seq_len(nrow(SUBGROUPS))) {
  cd <- SUBGROUPS$code[i]
  n  <- sum(eval(SUBGROUPS$rule[[i]], meta))
  rows[[length(rows)+1]] <- c(
    SUBGROUPS$label[i], as.character(n),
    mn(ga[cd,"Pattern_16"]), mn(pma[cd,"Pattern_16"]),   # CMAP_IGL_IMM (H-16)
    mn(ga[cd,"Pattern_17"]), mn(pma[cd,"Pattern_17"]),   # CMAP_IGL     (H-17)
    mn(ga[cd,"Pattern_27"]), mn(pma[cd,"Pattern_27"]))   # CMAP_PCL     (H-27)
}
S9 <- as.data.frame(do.call(rbind, rows), stringsAsFactors = FALSE)
names(S9) <- c("Subjects included","n","GA","PMA","GA ","PMA ","GA  ","PMA  ")

kable(S9, align = c("l","c","c","c","c","c","c","c"),
      caption = "Table S9. Correlation of CMAP activity with gestational and post-menstrual age within clinical subgroups.") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE) %>%
  add_header_above(c(" " = 2, "CMAP_IGL_IMM" = 2, "CMAP_IGL" = 2, "CMAP_PCL" = 2))
```

# Caption / abbreviations

```{r caption, echo=FALSE, results="asis"}
cat("Spearman rank-correlation coefficients (ρ) between subject-level CMAP activity and gestational age at birth (GA) or post-menstrual age at death (PMA). The first row is the full cohort; each subsequent row is restricted to the subjects indicated. Coefficients are descriptive; no P values are reported.\n\n")
cat("Abbreviations: CMAP, cerebellar maturation-associated pattern; GA, gestational age at birth; NEC, necrotizing enterocolitis; PMA, post-menstrual age at death.")
```

# Session Info

```{r session}
sessionInfo()
```
