---
title: "Table S5 — Gestational age group × post-menstrual age interaction in coronal cerebellar ultrasound growth, adjusted"
subtitle: "Single-covariate sensitivity: does the PMA × GA-group interaction survive adjusting for one clinical factor at a time?"
author: "Kratimenos Lab"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
suppressMessages({library(dplyr); library(tibble); library(knitr); library(kableExtra); library(readxl)})

find_data_dir <- function(sub = ".", max_up = 6) {
  prefix <- "."
  for (i in 0:max_up) { cand <- file.path(prefix, "Supplementary_Datasets", sub); if (dir.exists(cand)) return(cand)
    prefix <- file.path(prefix, "..") }
  stop("Supplementary_Datasets/", sub, " not found above ", normalizePath("."))
}
SRC  <- find_data_dir()
WB   <- file.path(SRC, "Data_S1_InVivo_Cohort.xlsx")   # consolidated: one sheet per former CSV
norm <- function(s) sub("^x(?=[a-z])", "", gsub("[^a-z0-9]", "", tolower(s)), perl = TRUE)
```

# Overview

Longitudinal coronal cerebellar ultrasound growth was modelled by linear mixed-effects regression with
a random intercept and post-menstrual-age (PMA) slope per infant. For each clinical factor, a base model
(*PMA + gestational-age (GA) group + factor*) was compared with a model additionally including the
*PMA × GA-group* interaction by likelihood-ratio test. The table reports, per factor, the χ² statistic,
the *P* value, and the number of scans (infants). The interaction is not retained under any single
adjustment (all *P* > 0.05).

# Load model summaries

```{r load}
us <- as.data.frame(read_excel(WB, sheet = "4 Coronal US growth vs GA"))
us$key <- norm(us$Covariate)
```

# Table S5

```{r tableS5, echo=FALSE}
s5_map <- tribble(
  ~label,                                                    ~src,
  "Maternal delivery indication",                            "Indication [Spont/Maternal/Unk]",
  "Surgical necrotizing enterocolitis (NEC)",                "NEC Surgical [Y/N/Unk]",
  "Cumulative burden of bronchopulmonary dysplasia (BPD)",   "BPD_score",
  "Cumulative burden of sepsis",                             "Sepsis_score",
  "Magnesium sulfate exposure",                              "Magnesium sulfate [Y/N/Unk]",
  "Caffeine exposure",                                       "Caffeine [Y/N/Unk]",
  "Glycemic control (glucose time in range, 70–180 mg/dL)", "Glucose TIR",
  "Cumulative procedural pain (PIPP)",                       "PIPP (cumulative)") %>%
  mutate(key = norm(src))

S5 <- s5_map %>% left_join(us, by = "key") %>%
  transmute(`Clinical factor adjusted for` = label,
            `χ²` = sprintf("%.2f", Chi.Square),
            `P` = sprintf("%.3f", P.Value),
            `Scans (infants)` = sprintf("%d (%d)", as.integer(N), as.integer(Subjects)))

kable(S5, align = c("l","c","c","c"),
      caption = "Table S5. Gestational age group × post-menstrual age interaction in coronal cerebellar ultrasound growth after adjustment for individual clinical factors (in vivo cohort).") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
```

# Abbreviations

```{r abbrev, echo=FALSE, results="asis"}
cat("Abbreviations: BPD, bronchopulmonary dysplasia; GA, gestational age at birth; NEC, necrotizing enterocolitis; PIPP, Premature Infant Pain Profile; PMA, post-menstrual age.")
```

# Session Info

```{r session}
sessionInfo()
```
