---
title: "Figure S11 (B, D) -- VGLUT1 / VGLUT2 positive-area fraction vs gestational age"
subtitle: "Fraction of the cerebellar molecular-layer ROI occupied by VGLUT1+ (B) and VGLUT2+ (D) signal"
author: ""
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, dpi = 300)
knitr::opts_knit$set(root.dir = normalizePath("."))
library(readxl); library(dplyr); library(readr); library(ggplot2)

# --- Location-agnostic resolvers (RELATIVE paths only) ---
find_data <- function(fname, subdir = NULL, max_up = 6) {
 prefix <- "."
 for (i in 0:max_up) {
 cand <- if (is.null(subdir)) file.path(prefix, "Supplementary_Datasets", fname) else file.path(prefix, "Supplementary_Datasets", subdir, fname)
 if (file.exists(cand)) return(cand)
 prefix <- file.path(prefix, "..")
 }
 stop(fname, " not found under any Supplementary_Datasets/ folder above ", normalizePath("."))
}
find_dir <- function(subdir, max_up = 6) {
 prefix <- "."
 for (i in 0:max_up) { cand <- file.path(prefix, "Supplementary_Datasets", subdir); if (dir.exists(cand)) return(cand); prefix <- file.path(prefix, "..") }
 stop("Supplementary_Datasets/", subdir, " not found above ", normalizePath("."))
}

VGLUT1_COL <- "#EE6677" # red (VGLUT1 channel)
VGLUT2_COL <- "#AA3377" # magenta/purple (VGLUT2 channel, matches the deck)
rlab <- function(r) sprintf('italic(r) == "%.3f"', r)
```

# Overview

Immunofluorescence quantification in the cerebellar molecular layer (ML): the **fraction of the ML ROI
occupied by VGLUT+ signal**, per subject, vs gestational age at birth. 8 post-mortem cerebella
(7 preterm, 1 term). Panels **A/C** (confocal images) are microscopy assets and are not generated here.

* **B** — VGLUT1+ area / total ML area (`Total_Cluster_Area / Cell_Area_um2`, subject-mean of 5 images).
* **D** — VGLUT2+ area / total ML area (`vglut2_pos_area_um2 / area_um2`, whole-ML row, subject-mean).

# Panel B -- VGLUT1+ area / total ML area

```{r panelB, fig.width=3.6, fig.height=3.4}
b <- read_excel(find_data("Data_S5_InSitu_Validation.xlsx"), sheet = "VGLUT1_Image_Level") %>%
 mutate(Subject_ID = as.character(Subject_ID),
 pos_area_frac = Total_Cluster_Area / Cell_Area_um2) %>%
 group_by(Subject_ID, GA_weeks) %>%
 summarise(pos_area_frac = mean(pos_area_frac), .groups = "drop") %>%
 rename(GA = GA_weeks)
ctB <- cor.test(b$GA, b$pos_area_frac)

pB <- ggplot(b, aes(GA, pos_area_frac)) +
 geom_smooth(method = "lm", se = TRUE, colour = "black", fill = "grey70", alpha = 0.3, linewidth = 0.7) +
 geom_point(colour = VGLUT1_COL, size = 3) +
 annotate("text", parse = TRUE, x = -Inf, y = Inf, hjust = -0.15, vjust = 1.6,
 label = rlab(ctB$estimate), size = 3.4, colour = "grey20") +
 labs(x = "Gestational age at birth (weeks)", y = "vGLUT1+ area / total ML area") +
 theme_classic(base_size = 10) +
 theme(axis.title = element_text(size = 9), axis.text = element_text(size = 8))
print(pB)
ggsave("Figure_S11B_VGLUT1_areafrac.pdf", pB, width = 3.6, height = 3.4, device = cairo_pdf, bg = "white")
cat(sprintf("VGLUT1: n=%d r=%.3f P=%.3f\n", nrow(b), ctB$estimate, ctB$p.value))
```

# Panel D -- VGLUT2+ area / total ML area

```{r panelD, fig.width=3.6, fig.height=3.4}
img <- read_excel(find_data("Data_S5_InSitu_Validation.xlsx"), sheet = "VGLUT2_Image_Level") %>%
 mutate(Study_ID = as.character(Subject_ID)) %>%
 filter(region == "ML") %>%
 mutate(pos_area_frac = vglut2_pos_area_um2 / area_um2)

cohort <- read_excel(find_data("Data_S2_Postmortem_Histology.xlsx"), sheet = "Postmortem_Cohort") %>%
 transmute(Study_ID = as.character(Paper_ID),
 Group = factor(Group, levels = c("Preterm", "Term")), GA = as.numeric(GA_wk))

d <- img %>% left_join(cohort, by = "Study_ID") %>%
 filter(Group == "Preterm" | Study_ID == "Term 2") %>%
 mutate(failed = mean_punctum_um2 > 100 | pct_positive_area > 90 | n_puncta < 10) %>%
 filter(!failed) %>%
 group_by(Study_ID, GA) %>%
 summarise(pos_area_frac = mean(pos_area_frac), .groups = "drop")
ctD <- cor.test(d$GA, d$pos_area_frac)

pD <- ggplot(d, aes(GA, pos_area_frac)) +
 geom_smooth(method = "lm", se = TRUE, colour = "black", fill = "grey70", alpha = 0.3, linewidth = 0.7) +
 geom_point(colour = VGLUT2_COL, size = 3) +
 annotate("text", parse = TRUE, x = -Inf, y = Inf, hjust = -0.15, vjust = 1.6,
 label = rlab(ctD$estimate), size = 3.4, colour = "grey20") +
 labs(x = "Gestational age at birth (weeks)", y = "vGLUT2+ area / total ML area") +
 theme_classic(base_size = 10) +
 theme(axis.title = element_text(size = 9), axis.text = element_text(size = 8))
print(pD)
ggsave("Figure_S11D_VGLUT2_areafrac.pdf", pD, width = 3.6, height = 3.4, device = cairo_pdf, bg = "white")
cat(sprintf("VGLUT2: n=%d r=%.3f P=%.3f\n", nrow(d), ctD$estimate, ctD$p.value))
```

# Statistics

```{r stats, echo=FALSE}
library(knitr); library(kableExtra)
tibble(
 Panel = c("B VGLUT1", "D VGLUT2"),
 n = c(nrow(b), nrow(d)),
 `Pearson r` = sprintf("%.3f", c(ctB$estimate, ctD$estimate)),
 `95% CI` = c(sprintf("[%.3f, %.3f]", ctB$conf.int[1], ctB$conf.int[2]),
 sprintf("[%.3f, %.3f]", ctD$conf.int[1], ctD$conf.int[2])),
 P = sprintf("%.3f", c(ctB$p.value, ctD$p.value))
) %>% kable(caption = "VGLUT+ area / total ML area vs GA (subject-level, n = 8)") %>%
 kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
```

# Provenance & data

* **B** source: VGLUT1 image-level quantification; data `Data_S5_InSitu_Validation.xlsx` sheet `VGLUT1_Image_Level`.
* **D** source: VGLUT2 per-image quantification; data = `Data_S5_InSitu_Validation.xlsx` sheet `VGLUT2_Image_Level` (per-image whole-ML rows) linked to the postmortem cohort key (`Data_S2_Postmortem_Histology.xlsx`, sheet `Postmortem_Cohort`); whole-ML QC + single-term (Term 2) selection.
* **Deck restyle:** r-only annotation (no title/P); VGLUT2 recoloured to **magenta #AA3377** (source used blue #4477AA) to match the deck's VGLUT2 channel; VGLUT1 red #EE6677.

# Session Info

```{r session}
sessionInfo()
```
