---
title: "Figure S13 (C, F, I) -- cell-cycle / proliferation markers"
subtitle: "EGL (C) & IGL (F) RNAScope H-scores + PHH3 nuclear cluster fraction (I) vs gestational age"
author: ""
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, dpi = 300)
knitr::opts_knit$set(root.dir = normalizePath("."))
library(readxl); library(dplyr); library(tidyr); library(ggplot2)

# --- Location-agnostic dataset finder (RELATIVE paths only) ---
find_dataset <- function(names, subdir = "Supplementary_Datasets", max_up = 6) {
 if (is.character(names) && length(names) == 1) names <- c(names)
 prefix <- "."
 for (i in 0:max_up) {
 for (nm in names) { rel <- file.path(prefix, subdir, nm); if (file.exists(rel)) return(rel) }
 prefix <- file.path(prefix, "..")
 }
 stop("Dataset not found under a 'Supplementary_Datasets/' folder: ", paste(names, collapse = " / "))
}

# marker colours (IF/RNAScope channels)
COL <- c(MCM2 = "#68BB3D", PCNA = "#DE33AB", CCNB1 = "#A4AB37", TOP2A = "#E03326")
PHH3_COL <- "#EE6677"
```

# Overview

Proliferation / cell-cycle markers in the developing cerebellum vs gestational age at birth.
Panels **A/B** (EGL schematic + IF), **D/E** (IGL), **G/H** (PC) are microscopy assets — not generated here.

| Panel | Layer | Markers | Metric |
|-------|-------|---------|--------|
| **C** | EGL | MCM2, PCNA, CCNB1, TOP2A | RNAScope H-score per ROI, subject mean ± SEM (descriptive; no correlation coefficient shown) |
| **F** | IGL | MCM2, CCNB1, TOP2A | RNAScope H-score per ROI, subject mean ± SEM (descriptive; no correlation coefficient shown) |
| **I** | PC | PHH3 | nuclear cluster fraction (subject-level), linear fit |

# Panels C & F -- RNAScope H-scores (EGL, IGL)

```{r hscore-data}
hscore <- read_excel(find_dataset("Data_S5_InSitu_Validation.xlsx"), sheet = "H_Score_ROI_CellCycle")

# single-marker: jittered ROI points + subject mean±SEM errorbar/line. No correlation coefficient is
# annotated: with n = 4 subjects, ROIs are not independent replicates, so trends are shown descriptively.
build_hscore <- function(dat, gene_name, cell, col) {
 pd <- dat %>% filter(Gene == gene_name, Cell_Type == cell)
 rd <- pd %>% group_by(GA) %>%
 summarise(mean_H = mean(H_score, na.rm = TRUE),
 se_H = sd(H_score, na.rm = TRUE) / sqrt(n()), .groups = "drop") %>%
 mutate(lo = pmax(mean_H - se_H, 0), hi = mean_H + se_H)
 ggplot() +
 geom_jitter(data = pd, aes(GA, H_score), color = col, size = 1.5, alpha = 0.4, width = 0.3) +
 geom_errorbar(data = rd, aes(GA, ymin = lo, ymax = hi), color = col, width = 1.5, linewidth = 0.6) +
 geom_line(data = rd, aes(GA, mean_H), color = col, linewidth = 0.8) +
 geom_point(data = rd, aes(GA, mean_H), color = col, size = 2.5) +
 scale_x_continuous(breaks = c(23, 28, 34.7, 40), labels = c("23", "28", "34.7", "40")) +
 labs(title = gene_name, x = "Gestational age at birth (weeks)", y = "H-score") +
 theme_classic(base_size = 11) +
 theme(plot.title = element_text(size = 11, hjust = 0.5),
 axis.title = element_text(size = 9), axis.text = element_text(size = 8))
}
save_panel <- function(p, stem, w = 3.4, h = 3.1) {
 ggsave(paste0(stem, ".pdf"), p, width = w, height = h, device = cairo_pdf, bg = "white")
}
```

## Panel C -- EGL (MCM2, PCNA, CCNB1, TOP2A)

```{r panelC, fig.width=3.4, fig.height=3.1}
egl <- c("MCM2", "PCNA", "CCNB1", "TOP2A")
for (i in seq_along(egl)) {
 p <- build_hscore(hscore, egl[i], "EGL", COL[[egl[i]]])
 print(p); save_panel(p, sprintf("Figure_S13C-%d_%s_EGL", i, egl[i]))
}
```

## Panel F -- IGL (MCM2, CCNB1, TOP2A)

```{r panelF, fig.width=3.4, fig.height=3.1}
igl <- c("MCM2", "CCNB1", "TOP2A")
for (i in seq_along(igl)) {
 p <- build_hscore(hscore, igl[i], "IGL", COL[[igl[i]]])
 print(p); save_panel(p, sprintf("Figure_S13F-%d_%s_IGL", i, igl[i]))
}
```

# Panel I -- PHH3 nuclear cluster fraction (PC)

```{r phh3-pipeline}
df <- read_excel(find_dataset("Data_S5_InSitu_Validation.xlsx"), sheet = "CyclinB_PH3_Cell_Level")

# Step 1: drop all-NaN PHH3 rows; 3-MAD outlier on detection events
key_cols <- c("PH3_Num_spots_estimated", "PH3_Num_single_spots", "PH3_Num_clusters")
df <- df %>% filter(rowSums(!is.na(select(., all_of(key_cols)))) > 0)
mad_flag <- function(x) {
 ok <- !is.na(x); out <- rep(FALSE, length(x))
 if (sum(ok) < 3) return(out)
 m <- median(x[ok]); s <- mad(x[ok]); if (s == 0) return(out)
 out[ok] <- abs(x[ok] - m) > 3 * s; out
}
df <- df %>% mutate(det_events = PH3_Num_single_spots + PH3_Num_clusters,
 outlier = mad_flag(det_events)) %>%
 filter(!outlier)

# Step 2/3: nuclear cluster fraction = nuclear clusters / total nuclear puncta
df <- df %>% mutate(Nuclear_cluster_fraction =
 ifelse(Nuclear_puncta > 0, Nuclear_det_clusters / Nuclear_puncta, NA_real_))

# aggregate: image-level then subject-level (n = 8)
image_level <- df %>% group_by(Subject_ID, Image_Num, GA_weeks, Group) %>%
 summarise(f = mean(Nuclear_cluster_fraction, na.rm = TRUE), .groups = "drop")
subj <- image_level %>% group_by(Subject_ID, GA_weeks, Group) %>%
 summarise(f = mean(f, na.rm = TRUE), .groups = "drop")
```

```{r panelI, fig.width=3.4, fig.height=3.2}
pI <- ggplot(subj, aes(GA_weeks, f)) +
 geom_smooth(method = "lm", se = FALSE, colour = PHH3_COL, linewidth = 0.8) +
 geom_point(colour = PHH3_COL, size = 3) +
 labs(title = "PHH3", x = "Gestational age at birth (weeks)", y = "Nuclear cluster fraction") +
 theme_classic(base_size = 11) +
 theme(plot.title = element_text(size = 11, hjust = 0.5),
 axis.title = element_text(size = 9), axis.text = element_text(size = 8))
print(pI)
save_panel(pI, "Figure_S13I_PHH3_Nuclear_Cluster", w = 3.4, h = 3.2)
```

# Statistics

```{r stats, echo=FALSE}
library(knitr); library(kableExtra)
spec <- rbind(
 data.frame(Panel = "C", Layer = "EGL", Marker = c("MCM2","PCNA","CCNB1","TOP2A")),
 data.frame(Panel = "F", Layer = "IGL", Marker = c("MCM2","CCNB1","TOP2A"))
)
spec$N_subjects <- NA_integer_; spec$N_ROI <- NA_integer_
for (i in seq_len(nrow(spec))) {
 pd <- hscore %>% filter(Gene == spec$Marker[i], Cell_Type == spec$Layer[i])
 spec$N_subjects[i] <- n_distinct(pd$Subject_ID); spec$N_ROI[i] <- nrow(pd)
}
ph <- cor.test(subj$GA_weeks, subj$f)
spec$r <- ""
spec <- rbind(spec, data.frame(Panel = "I", Layer = "PC", Marker = "PHH3",
 N_subjects = nrow(subj), N_ROI = NA_integer_, r = sprintf("%+.2f", ph$estimate)))
kable(spec, caption = "Panels C/F: subjects and ROIs per marker (descriptive, no correlation reported). Panel I: Pearson r (subject-level) vs GA") %>%
 kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
```

# Provenance

Source: H-score quantification (panels C/F) + PHH3 immunostaining quantification (panel I).
Data: `Data_S5_InSitu_Validation.xlsx` sheet `H_Score_ROI_CellCycle`; `Data_S5_InSitu_Validation.xlsx` sheet `CyclinB_PH3_Cell_Level`.
**Panel numbering (cell-cycle figure, S12):**
panel **C = EGL**, **F = IGL**, **I = PHH3** (EGL/IGL labels corrected relative to an earlier version).
Deck edits: x-axis "Gestational age at birth (weeks)"; plain titles. Correlation annotations removed from C/F (ROI-level
coefficients are not reported; n = 4 subjects).
Marker colours MCM2 #68BB3D / PCNA #DE33AB / CCNB1 #A4AB37 / TOP2A #E03326; PHH3 #EE6677.

# Session Info

```{r session}
sessionInfo()
```
