---
title: "Figure S16 -- Gestational-age effect on cerebellar layer architecture (covariate robustness)"
subtitle: "Post-mortem cohort: GA effect on each cerebellar layer after adjusting for one covariate at a time"
author: ""
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, dpi = 150)
library(dplyr); library(tidyr); library(ggplot2); library(cowplot); library(knitr); library(kableExtra); library(readxl)

# --- Location-agnostic resolver: walk up to Supplementary_Datasets/Covariate_Robustness/ ---
find_data_dir <- function(subdir = ".", max_up = 6) {
 prefix <- "."
 for (i in 0:max_up) {
 cand <- file.path(prefix, "Supplementary_Datasets", subdir)
 if (dir.exists(cand)) return(cand)
 prefix <- file.path(prefix, "..")
 }
 stop("Supplementary_Datasets/", subdir, " not found above ", normalizePath("."))
}
SRC <- find_data_dir()
WB  <- file.path(SRC, "Data_S2_Postmortem_Histology.xlsx")   # consolidated: one sheet per former CSV
ALPHA <- 0.05

# Deck palette: retained = black, not retained = magenta.
COL_RETAIN <- "#DE1567"; COL_NOTRETAIN <- "#8A8A8A" # deck 2026-08: retained = pink, not retained = grey
LVL <- c("Association retained (P < 0.05)", "Not retained after adjustment")
```

# Overview

Post-mortem histology cohort: for each cerebellar layer, the gestational-age effect (per week, 95% CI)
after adjusting for each selected clinical covariate **one at a time**.
*Black* = association retained (`P < 0.05`); *magenta* = not retained after adjustment.
Panels: **A** EGL surface area · **B** ML surface area · **C** IGL surface area · **D** PC count.
Each panel is exported as a separate file (never combined in code).

# Data

```{r data}
norm <- function(s) sub("^x(?=[a-z])", "", gsub("[^a-z0-9]", "", tolower(s)), perl = TRUE)
pm <- as.data.frame(read_excel(WB, sheet = "5 Layers vs GA"))
pm$key <- norm(pm$Covariate)

# Curated covariates in the deck's top->bottom order
CURATED <- tribble(
 ~display, ~source,
 "Maternal delivery indication", "Indication [Spont/Maternal/Unk]",
 "Surgical NEC", "NEC Surgical [Y/N/Unk]",
 "Cumulative burden of BPD", "BPD_score",
 "Cumulative burden of sepsis", "Sepsis_score",
 "Glycemic control", "Glucose TIR",
 "Cumulative procedural pain", "PIPP (cumulative)",
 "Magnesium sulfate", "Magnesium sulfate [Y/N/Unk]",
 "Caffeine", "Caffeine [Y/N/Unk]") %>%
 mutate(key = norm(source))
row_levels <- rev(CURATED$display) # ggplot(): first level at bottom -> Delivery ends up on top

LAYERS <- tribble(
 ~outcome, ~title,
 "EGL_normalized", "EGL surface area",
 "IGL_normalized", "IGL surface area",
 "ML_normalized", "ML surface area",
 "PC_count", "PC count")
```

# Forest builder + shared legend

```{r builder}
layer_forest <- function(outcome, title) {
 d <- pm %>% filter(Outcome == outcome)
 df <- CURATED %>% left_join(d, by = "key") %>%
 transmute(row = factor(display, levels = row_levels),
 est = GA_beta, lo = CI_low, hi = CI_high,
 status = factor(ifelse(!is.na(P_GA) & P_GA < ALPHA, LVL[1], LVL[2]), levels = LVL))
 ggplot(df, aes(y = row)) +
 geom_vline(xintercept = 0, linetype = "22", colour = "#9A9A9A") +
 geom_linerange(aes(xmin = lo, xmax = hi, colour = status), linewidth = 0.8, show.legend = FALSE) +
 geom_point(aes(x = est, fill = status), shape = 21, size = 2.0, colour = "white", stroke = 0.7,
 show.legend = FALSE) +
 scale_colour_manual(values = setNames(c(COL_RETAIN, COL_NOTRETAIN), LVL), drop = FALSE) +
 scale_fill_manual(values = setNames(c(COL_RETAIN, COL_NOTRETAIN), LVL), drop = FALSE) +
 labs(title = title, x = "GA effect (per week, 95% CI)", y = NULL) +
 theme_minimal(base_size = 10) +
 theme(panel.grid.major.y = element_blank(), panel.grid.minor = element_blank(),
 panel.grid.major.x = element_line(colour = "#EFEFEF"),
 panel.border = element_rect(colour = "#B0B0B0", fill = NA, linewidth = 0.3),
 plot.title = element_text(hjust = 0.5, size = 10, colour = "#333333"),
 axis.text.y = element_text(colour = "#222222", size = 8),
 axis.text.x = element_text(size = 7.5, colour = "#333333"),
 axis.title.x = element_text(size = 8, colour = "#333333"),
 plot.margin = margin(6, 10, 4, 4))
}

save_panel <- function(p, stem, w = 4.4, h = 2.7) {
 ggsave(paste0(stem, ".pdf"), p, width = w, height = h, device = cairo_pdf, bg = "white")
}

# shared legend (points, black = retained / magenta = not retained)
leg_src <- ggplot(data.frame(x = 1:2, y = 1:2, status = factor(LVL, levels = LVL)),
 aes(x, y, fill = status)) +
 geom_point(shape = 21, size = 2.6, colour = "white", stroke = 0.7) +
 scale_fill_manual(values = setNames(c(COL_RETAIN, COL_NOTRETAIN), LVL), name = NULL) +
 theme_void() +
 theme(legend.position = "bottom", legend.text = element_text(size = 9),
 legend.key = element_blank())
shared_legend <- cowplot::get_legend(leg_src)
```

# Panels

```{r panels, fig.width=4.4, fig.height=2.7, echo=FALSE, results="asis"}
stems <- c("Figure_S16A_EGL_surface_area", "Figure_S16B_IGL_surface_area",
 "Figure_S16C_ML_surface_area", "Figure_S16D_PC_count")
for (i in seq_len(nrow(LAYERS))) {
 p <- layer_forest(LAYERS$outcome[i], LAYERS$title[i])
 save_panel(p, stems[i])
 print(p)
}
# shared legend as its own file
ggsave("Figure_S16_legend.pdf", ggdraw(shared_legend), width = 5.5, height = 0.4, device = cairo_pdf, bg = "white")
```

# Table S14 — GA effect on cerebellar layer architecture, adjusted for individual clinical factors

Rendered inline from the same model summaries used by the forest panels above (values match the
submitted Table S14). Each covariate is shown once across its four layer rows; *n* is the model
sample size for that covariate. Note the table lists **sepsis before BPD** (the submitted-table
order), whereas the figure panels use the deck order.

```{r tableS14, echo=FALSE}
s14_cov <- tribble(
  ~label,                         ~src,
  "Maternal delivery indication", "Indication [Spont/Maternal/Unk]",
  "Surgical NEC",                 "NEC Surgical [Y/N/Unk]",
  "Cumulative burden of sepsis",  "Sepsis_score",
  "Cumulative burden of BPD",     "BPD_score",
  "Glycemic control",             "Glucose TIR",
  "Cumulative procedural pain",   "PIPP (cumulative)",
  "Magnesium sulfate",            "Magnesium sulfate [Y/N/Unk]",
  "Caffeine",                     "Caffeine [Y/N/Unk]") %>%
  mutate(cov_i = row_number(), key = norm(src))

s14_lay <- tribble(
  ~outcome,          ~layer,
  "EGL_normalized",  "EGL area",
  "IGL_normalized",  "IGL area",
  "ML_normalized",   "ML area",
  "PC_count",        "PC density") %>%
  mutate(lay_i = row_number())

S14 <- tidyr::crossing(cov_i = s14_cov$cov_i, lay_i = s14_lay$lay_i) %>%
  left_join(s14_cov, by = "cov_i") %>%
  left_join(s14_lay, by = "lay_i") %>%
  left_join(pm, by = c("key" = "key", "outcome" = "Outcome")) %>%
  arrange(cov_i, lay_i) %>%
  group_by(cov_i) %>%
  mutate(n_show = ifelse(row_number() == 1, as.character(N), "")) %>%
  ungroup() %>%
  transmute(`Clinical factor adjusted for` = label,
            `Layer` = layer,
            `GA β per week` = sprintf("%.2f", GA_beta),
            `95% CI` = sprintf("%.2f to %.2f", CI_low, CI_high),
            `P` = ifelse(P_GA < 0.001, "<0.001", sprintf("%.3f", P_GA)),
            `n` = n_show)

kable(S14, align = c("l","l","r","c","r","r"),
      caption = "Table S14. Gestational age effect on cerebellar layer architecture after adjustment for individual clinical factors (postmortem neuropathology cohort).") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE) %>%
  collapse_rows(columns = 1, valign = "top")
```

# Session Info

```{r session}
sessionInfo()
```
