---
title: "Figure S16 -- External granular layer (EGL) cell density"
subtitle: "GCP density (detections / layer area) vs gestational age at birth (A) and post-menstrual age at death (B)"
author: ""
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center", dpi = 300)
knitr::opts_knit$set(root.dir = normalizePath("."))
library(tidyverse); library(readxl)

# --- Location-agnostic resolver: walk up to Supplementary_Datasets/EGL_Density/, return RELATIVE path ---
find_egl_file <- function(fname, subdir = ".", max_up = 6) {
 prefix <- "."
 for (i in 0:max_up) {
 cand <- file.path(prefix, "Supplementary_Datasets", subdir, fname)
 if (file.exists(cand)) return(cand)
 prefix <- file.path(prefix, "..")
 }
 stop(fname, " not found under any Supplementary_Datasets/", subdir, "/ above ", normalizePath("."))
}
DAT <- find_egl_file("Dataset_EGL_Density.xlsx") # per-detection counts + areas
ORI <- find_egl_file("Dataset_Cerebellar_Layers.xlsx", subdir = ".") # GA / PMA-at-death / Group (identical to Fig 3)
```

# Overview

External granule layer (EGL) **cell density** (granule-cell precursors, GCPs) for the post-mortem
histopathology cohort, in the **Figure 3 design** (same palette, point/line sizes, axis ranges, and on-figure
statistics). Density = pooled detections ÷ pooled layer area per subject (one value/subject). Axes and
preterm/term grouping are taken from the Figure 3 dataset so the cohort matches exactly.

* **A** — EGL density vs **gestational age at birth** (single OLS trend across all subjects; dashed line at 37 wk).
* **B** — EGL density vs **post-menstrual age at death** (separate preterm/term trends with 95% CI ribbons;
 per-group slope β [95% CI] and slope-difference P on-figure).

```{r data, include=FALSE}
# per-subject EGL density (detections per mm^2)
egl_density <- read_excel(DAT, sheet = "EGL") %>%
 mutate(Subject_ID = as.character(Paper_ID)) %>%
 group_by(Subject_ID) %>%
 summarise(num = sum(`Num EGL`, na.rm = TRUE),
 area = sum(`Area µm^2`, na.rm = TRUE), .groups = "drop") %>%
 mutate(EGL_density = num / area * 1e6)

# GA / chronological (PMA-at-death) / Group from the Figure 3 dataset -> identical grouping & order
axes <- read_excel(ORI, sheet = "Cerebellar_Layers") %>%
 transmute(Subject_ID = as.character(Subject_ID), GA, Chronological_age, Group) %>%
 distinct(Subject_ID, .keep_all = TRUE)

data_egl <- egl_density %>% inner_join(axes, by = "Subject_ID")
```

```{r styling, include=FALSE}
# ---- Figure 3 deck-matched styling (ported verbatim) --------------------------
color_preterm <- "#030403"; color_term <- "#DE1567" # black preterm / crimson term ( 2026-08 palette)
line_preterm <- "#030403"; line_term <- "#DE1567" # trend-line colours ( 2026-08 palette)
line_thickness <- 0.8; point_size <- 1.5; point_alpha <- 0.7

test_slope_difference <- function(x1, y1, x2, y2) {
 n1 <- sum(!is.na(x1) & !is.na(y1)); n2 <- sum(!is.na(x2) & !is.na(y2))
 if (n1 < 3 || n2 < 3) return(list(p_value = NA))
 v1 <- !is.na(x1)&!is.na(y1); v2 <- !is.na(x2)&!is.na(y2)
 x1<-x1[v1];y1<-y1[v1];x2<-x2[v2];y2<-y2[v2]
 m1<-lm(y1~x1); m2<-lm(y2~x2)
 s1<-coef(m1)[2]; e1<-summary(m1)$coefficients[2,2]
 s2<-coef(m2)[2]; e2<-summary(m2)$coefficients[2,2]
 t<-(s1-s2)/sqrt(e1^2+e2^2); df<-n1+n2-4
 list(p_value = 2*pt(abs(t), df, lower.tail = FALSE))
}

fmt_p <- function(p) { # for "Slope P" (plotmath, no asterisks)
 if (is.na(p)) return("italic(P) == 'NA'")
 if (p < 0.001) return("italic(P) < \"0.001\"")
 paste0('italic(P) == "', sprintf("%.3f", p), '"')
}
fmt_p_beta <- function(p) {
 if (is.na(p)) return("italic(P) == 'NA'")
 if (p < 0.0001) return("italic(P) < \"0.0001\"")
 if (p < 0.001) return("italic(P) < \"0.001\"")
 paste0('italic(P) == "', sprintf("%.3f", p), '"')
}

# Returns the two panels SEPARATELY (never combined in code).
egl_panels <- function(df, y_lab) {
 df <- df %>% filter(!is.na(EGL_density), !is.na(GA), !is.na(Chronological_age))
 preterm_df <- df %>% filter(Group == "Preterm"); term_df <- df %>% filter(Group == "Term")

 s_pre <- summary(lm(preterm_df$EGL_density ~ preterm_df$Chronological_age))$coefficients
 s_ter <- summary(lm(term_df$EGL_density ~ term_df$Chronological_age))$coefficients
 slope_preterm_ca<-s_pre[2,1]; se_pre<-s_pre[2,2]; p_pre<-s_pre[2,4]
 slope_term_ca <-s_ter[2,1]; se_ter<-s_ter[2,2]; p_ter<-s_ter[2,4]
 ci_lo_preterm<-slope_preterm_ca-1.96*se_pre; ci_hi_preterm<-slope_preterm_ca+1.96*se_pre
 ci_lo_term <-slope_term_ca -1.96*se_ter; ci_hi_term <-slope_term_ca +1.96*se_ter
 p_slope_all_ga <- summary(lm(df$EGL_density ~ df$GA))$coefficients[2,4]
 slope_test <- test_slope_difference(preterm_df$Chronological_age, preterm_df$EGL_density,
 term_df$Chronological_age, term_df$EGL_density)

 y_min<-min(df$EGL_density); y_max<-max(df$EGL_density); y_range<-y_max-y_min
 y_limits<-c(y_min - y_range*0.08, y_max + y_range*0.22)

 base_theme <- theme_bw(base_size = 10) +
 theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
 panel.border = element_rect(color = "gray30", linewidth = 0.5),
 axis.title.x = element_text(size = 7, color = "gray30"),
 axis.title.y = element_text(size = 7.5, color = "gray30"),
 axis.text = element_text(size = 6.5, color = "gray30"),
 legend.position = "none", plot.margin = margin(8, 12, 4, 6))

 # ---- A: GA at birth (single OLS across all subjects; gray line) ----
 fit_all <- lm(EGL_density ~ GA, data = df)
 ga_seq <- seq(min(df$GA), max(df$GA), length.out = 200)
 pred <- predict(fit_all, newdata = data.frame(GA = ga_seq), interval = "confidence")
 pred_df <- data.frame(GA = ga_seq, fit = pred[,"fit"], lwr = pred[,"lwr"], upr = pred[,"upr"])

 p_ga <- ggplot(df, aes(GA, EGL_density, color = Group)) +
 geom_point(size = point_size, alpha = point_alpha) +
 geom_ribbon(data = pred_df, inherit.aes = FALSE, aes(GA, ymin = lwr, ymax = upr),
 fill = "gray60", alpha = 0.15) +
 geom_line(data = pred_df, inherit.aes = FALSE, aes(GA, fit),
 color = "gray30", linewidth = line_thickness) +
 geom_vline(xintercept = 37, linetype = "dashed", color = "gray70", linewidth = 0.5) +
 scale_color_manual(values = c(Preterm = color_preterm, Term = color_term)) +
 scale_x_continuous(limits = c(20, 45), breaks = seq(20, 45, 5)) +
 scale_y_continuous(limits = y_limits) +
 labs(x = "Gestational age at birth (weeks)", y = y_lab) + base_theme +
 annotate("text", parse = TRUE, x = 44, y = y_max + y_range*0.16,
 label = paste0('"Slope"~~', fmt_p(p_slope_all_ga)),
 hjust = 1, size = 1.8, color = "gray30")

 # ---- B: PMA at death (per-group trends + CI; slope stats) ----
 txt_x <- 60; txt_y1 <- y_max + y_range*0.22; txt_y2 <- y_max + y_range*0.15; txt_y3 <- y_max + y_range*0.08
 p_ca <- ggplot(df, aes(Chronological_age, EGL_density)) +
 geom_point(aes(color = Group), size = point_size, alpha = point_alpha) +
 geom_smooth(data = preterm_df, aes(Chronological_age, EGL_density), method = "lm", se = TRUE,
 color = line_preterm, fill = line_preterm, alpha = 0.15, linewidth = line_thickness) +
 geom_smooth(data = term_df, aes(Chronological_age, EGL_density), method = "lm", se = TRUE,
 color = line_term, fill = line_term, alpha = 0.15, linewidth = line_thickness) +
 scale_color_manual(values = c(Preterm = color_preterm, Term = color_term)) +
 scale_x_continuous(limits = c(0, 60), breaks = seq(0, 60, 10)) +
 scale_y_continuous(limits = y_limits) +
 labs(x = "Chronological age (weeks)", y = y_lab) + base_theme +
 annotate("text", parse = TRUE, x = txt_x, y = txt_y1,
 label = paste0('"Slope"~~', fmt_p(slope_test$p_value)), hjust = 1, size = 1.8, color = "gray30") +
 annotate("text", parse = TRUE, x = txt_x, y = txt_y2,
 label = paste0("paste('Preterm: ', italic(beta), ' = ", sprintf("%.2f", slope_preterm_ca),
 " [", sprintf("%.2f", ci_lo_preterm), "–", sprintf("%.2f", ci_hi_preterm), "], ', ",
 fmt_p_beta(p_pre), ")"), hjust = 1, size = 1.4, color = line_preterm) +
 annotate("text", parse = TRUE, x = txt_x, y = txt_y3,
 label = paste0("paste('Term: ', italic(beta), ' = ", sprintf("%.2f", slope_term_ca),
 " [", sprintf("%.2f", ci_lo_term), "–", sprintf("%.2f", ci_hi_term), "], ', ",
 fmt_p_beta(p_ter), ")"), hjust = 1, size = 1.4, color = line_term)

 list(p_ga = p_ga, p_ca = p_ca)
}
```

# Figure S16A -- EGL density vs gestational age at birth

```{r figA, fig.width=2.2, fig.height=3.3, echo=FALSE}
Y_LAB <- expression(EGL~density~(GCPs/mm^2))
panels <- egl_panels(data_egl, Y_LAB)
print(panels$p_ga)
ggsave("Figure_S16A_EGL_vs_GAbirth.pdf", panels$p_ga, width = 2.2, height = 3.3, device = cairo_pdf)
```

# Figure S16B -- EGL density vs post-menstrual age at death

```{r figB, fig.width=2.2, fig.height=3.3, echo=FALSE}
print(panels$p_ca)
ggsave("Figure_S16B_EGL_vs_PMAdeath.pdf", panels$p_ca, width = 2.2, height = 3.3, device = cairo_pdf)
```

# Statistics

```{r stats, echo=FALSE}
d <- data_egl; pre <- d %>% filter(Group=="Preterm"); ter <- d %>% filter(Group=="Term")
tibble(
 Metric = c("N total","N preterm","N term",
 "GA-slope P (all)","Slope-difference P (preterm vs term)"),
 Value = c(nrow(d), nrow(pre), nrow(ter),
 sprintf("%.3f", summary(lm(d$EGL_density ~ d$GA))$coefficients[2,4]),
 sprintf("%.3f", test_slope_difference(pre$Chronological_age, pre$EGL_density,
 ter$Chronological_age, ter$EGL_density)$p_value))
) %>% knitr::kable() %>% kableExtra::kable_styling(full_width = FALSE)
```

# Session Info

```{r session}
sessionInfo()
```
