---
title: "Figure S5C: Midsagittal Cerebellar Composite (Z-score) by GA Group at 28/33/35 wk PMA"
subtitle: "Cross-sectional ultrasound composite; pairwise Wilcoxon (was old Figure S4C)"
author: ""
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
 fig.width = 3.42, fig.height = 5.46, dpi = 300)
knitr::opts_knit$set(root.dir = normalizePath("."))

# --- Local dataset finder (RELATIVE paths only) ---------------------------
find_dataset <- function(names, subdir = "Supplementary_Datasets", max_up = 6) {
 if (is.character(names) && length(names) == 1) names <- c(names)
 prefix <- "."
 for (i in 0:max_up) {
 for (nm in names) { rel <- file.path(prefix, subdir, nm); if (file.exists(rel)) return(rel) }
 prefix <- file.path(prefix, "..")
 }
 stop("Dataset not found under a 'Supplementary_Datasets/' folder: ", paste(names, collapse = " / "))
}
```

# Overview

Supplement panel **Figure S5C** — the **midsagittal** counterpart of main **Figure 1E** (coronal), styled
identically. Midsagittal cerebellar composite (inverse-correlation weighted z-score of midsagittal AP, height,
area) across GA groups (EP / VP / LP) at 28, 33, 35 wk ± 2 PMA. One scan per patient (closest to target).
Pairwise two-sided **Wilcoxon rank-sum**. Was previously **Figure S4C**. Per the original S4 pipeline, the
midsagittal composite is computed on the coronal-valid subset.

```{r libs}
library(readxl); library(dplyr); library(ggplot2); library(grid)
library(knitr); library(kableExtra)

iqr_outlier <- function(x, k = 3) {
 q1 <- quantile(x, .25, na.rm = TRUE); q3 <- quantile(x, .75, na.rm = TRUE); iqr <- q3 - q1
 x < (q1 - k*iqr) | x > (q3 + k*iqr)
}
inv_corr_composite <- function(df, cols) {
 m <- cor(df %>% select(all_of(cols)) %>% filter(complete.cases(.)), use = "pairwise.complete.obs")
 w <- 1 - sapply(seq_along(cols), function(i) mean(abs(m[i, -i]))); w <- w / sum(w)
 Reduce(`+`, Map(function(cn, wt) wt * as.numeric(scale(df[[cn]])), cols, w))
}
```

# Data + midsagittal composite (reproduces old S4 pipeline)

```{r load-composite}
data_file <- find_dataset(c("Dataset_Clinical_MRI_US_ND.xlsx", "Dataset_S1_Clinical_MRI_US_ND.xlsx"))

us_data <- read_excel(data_file, sheet = "Ultrasound") %>%
 mutate(
 GA_Category = factor(case_when(
 GA_weeks < 28 ~ "Extremely Preterm",
 GA_weeks < 32 ~ "Very Preterm",
 TRUE ~ "Late Preterm"),
 levels = c("Extremely Preterm", "Very Preterm", "Late Preterm")),
 Study_ID = as.character(Study_ID),
 across(c(Coronal_Height, Coronal_Transverse, Coronal_Area,
 Midsag_AP, Midsag_Height, Midsag_Area), as.numeric)
 ) %>%
 filter(!is.na(PMA_weeks) & PMA_weeks >= 23 & PMA_weeks <= 45)

# CORONAL first (defines the analysis subset, matching original S4 pipeline)
us_data <- us_data %>%
 mutate(Coronal_Height = ifelse(iqr_outlier(Coronal_Height), NA, Coronal_Height),
 Coronal_Transverse = ifelse(iqr_outlier(Coronal_Transverse), NA, Coronal_Transverse),
 Coronal_Area = ifelse(iqr_outlier(Coronal_Area), NA, Coronal_Area))
us_data$US_Coronal_Composite <- inv_corr_composite(
 us_data, c("Coronal_Height","Coronal_Transverse","Coronal_Area"))
us_data <- us_data %>% filter(!is.na(US_Coronal_Composite))

# MIDSAGITTAL on the coronal-valid subset
us_data <- us_data %>%
 mutate(Midsag_AP = ifelse(iqr_outlier(Midsag_AP), NA, Midsag_AP),
 Midsag_Height = ifelse(iqr_outlier(Midsag_Height), NA, Midsag_Height),
 Midsag_Area = ifelse(iqr_outlier(Midsag_Area), NA, Midsag_Area))
us_data$US_Midsag_Composite <- inv_corr_composite(
 us_data, c("Midsag_AP","Midsag_Height","Midsag_Area"))
us_data <- us_data %>% filter(!is.na(US_Midsag_Composite))
```

# Figure S5C

```{r figure-s3d, fig.width=3.42, fig.height=5.46}
ga_short <- c("Extremely Preterm" = "EP", "Very Preterm" = "VP", "Late Preterm" = "LP")
point_cols <- c(EP = "#030403", VP = "#DE1567", LP = "#7BACDC") # ' updated palette

fmt_p_expr <- function(p) {
 if (is.na(p)) return("NA")
 if (p < 0.001) return("italic(P) < 0.001")
 paste0("italic(P) == '", sprintf("%.3f", p), "'")
}

# one scan/patient nearest a target PMA (± window); keep groups with n >= 2
get_tp <- function(data, target, window = 2) {
 data %>%
 filter(PMA_weeks >= target - window, PMA_weeks <= target + window) %>%
 group_by(Study_ID) %>% slice_min(abs(PMA_weeks - target), n = 1, with_ties = FALSE) %>%
 ungroup() %>%
 mutate(GA_Label = factor(ga_short[as.character(GA_Category)], levels = c("EP","VP","LP"))) %>%
    group_by(GA_Label) %>% filter(n() >= 2) %>% ungroup()
}

# --- Built like Figure 1E: single continuous box + dashed separators + top headers ---
grp_order <- c("EP", "VP", "LP")
grp_step <- 0.8
grp_gap <- 1.1
tp_labels <- c(`28` = "28 weeks\nPMA", `33` = "33 weeks\nPMA", `35` = "35 weeks\nPMA")

plot_df <- list(); bracket_df <- list(); header_df <- list(); sep_x <- c()
cursor <- 1; first <- TRUE
for (tp in names(tp_labels)) {
 d <- get_tp(us_data, as.numeric(tp))
 present <- intersect(grp_order, as.character(unique(droplevels(d$GA_Label))))
 xs <- cursor + (seq_along(present) - 1) * grp_step
 names(xs) <- present
 if (!first) sep_x <- c(sep_x, cursor - grp_gap / 2)
 d$xc <- xs[as.character(d$GA_Label)]
 plot_df[[tp]] <- d %>% transmute(GA_Label, y = US_Midsag_Composite, xc)
 header_df[[tp]] <- data.frame(xl = min(xs) - 0.45, xr = max(xs) + 0.45,
 xc = mean(range(xs)), text = tp_labels[tp])
 combos <- combn(present, 2, simplify = FALSE)
 combos <- combos[order(sapply(combos, function(cc) abs(xs[cc[2]] - xs[cc[1]])))]
 ymax <- max(d$US_Midsag_Composite, na.rm = TRUE); base <- ymax + 0.30; step <- 0.42
 for (i in seq_along(combos)) {
 cc <- combos[[i]]
 p <- wilcox.test(d$US_Midsag_Composite[d$GA_Label == cc[1]],
 d$US_Midsag_Composite[d$GA_Label == cc[2]])$p.value
 bracket_df[[paste(tp,i)]] <- data.frame(
 x = xs[cc[1]], xend = xs[cc[2]], y = base + step*(i-1),
 lx = mean(c(xs[cc[1]], xs[cc[2]])), label = fmt_p_expr(p))
 }
 cursor <- max(xs) + grp_gap; first <- FALSE
}
plot_df <- bind_rows(plot_df)
bracket_df <- bind_rows(bracket_df)
header_df <- bind_rows(header_df)
x_breaks <- plot_df %>% distinct(GA_Label, xc) %>% arrange(xc)

tick <- 0.10
top_bracket <- max(bracket_df$y) + 0.22
y_lo <- min(plot_df$y, na.rm = TRUE) - 0.20
x_left <- min(plot_df$xc) - 0.6
x_right <- max(plot_df$xc) + 0.5
box_top <- top_bracket + 0.05 # full box encloses data + brackets; headers sit above it
hdr_line_y <- box_top + 0.22 # underline: small gap above the box top
hdr_text_y <- hdr_line_y + 0.24 # header text above its underline

fig_s3d <- ggplot(plot_df, aes(x = xc, y = y)) +
 # dashed separators between timepoints (span the full box height)
 geom_segment(data = data.frame(x = sep_x),
 aes(x = x, xend = x, y = y_lo, yend = box_top),
 inherit.aes = FALSE, linetype = "dashed", colour = "grey70", linewidth = 0.3) +
 geom_boxplot(aes(group = xc), width = 0.5, fill = "white",
 colour = "gray30", linewidth = 0.45, outlier.shape = NA) +
 geom_jitter(aes(colour = GA_Label), width = 0.11, alpha = 1, size = 1.1) +
 # significance brackets
 geom_segment(data = bracket_df, aes(x = x, xend = xend, y = y, yend = y),
 inherit.aes = FALSE, linewidth = 0.35, colour = "gray20") +
 geom_segment(data = bracket_df, aes(x = x, xend = x, y = y - tick, yend = y),
 inherit.aes = FALSE, linewidth = 0.35, colour = "gray20") +
 geom_segment(data = bracket_df, aes(x = xend, xend = xend, y = y - tick, yend = y),
 inherit.aes = FALSE, linewidth = 0.35, colour = "gray20") +
 geom_text(data = bracket_df, aes(x = lx, y = y + tick*1.6, label = label),
 inherit.aes = FALSE, parse = TRUE, size = 3.0, colour = "gray20") +
 # timepoint headers (underlined) above the box
 geom_segment(data = header_df, aes(x = xl, xend = xr, y = hdr_line_y, yend = hdr_line_y),
 inherit.aes = FALSE, linewidth = 0.5, colour = "black") +
 geom_text(data = header_df, aes(x = xc, y = hdr_text_y, label = text),
 inherit.aes = FALSE, size = 3.1, colour = "black", lineheight = 0.9, vjust = 0) +
 scale_colour_manual(values = point_cols, guide = "none") +
 scale_x_continuous(breaks = x_breaks$xc, labels = x_breaks$GA_Label,
 limits = c(x_left, x_right), expand = c(0, 0)) +
 scale_y_continuous(breaks = c(-2, 0, 2)) +
 coord_cartesian(ylim = c(y_lo, box_top), expand = FALSE, clip = "off") +
 labs(x = NULL, y = "Midsagittal Composite (Z-score)") +
 theme_classic(base_size = 10) +
 theme(
 axis.text.x = element_text(size = 10),
 axis.text.y = element_text(size = 10),
 axis.title.y = element_text(size = 10),
 axis.line = element_blank(),
 panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5),
 axis.ticks = element_line(linewidth = 0.3, colour = "black"),
 panel.grid = element_blank(),
 plot.margin = margin(t = 58, r = 6, b = 3, l = 3)
 )

fig_s3d
ggsave("Figure_S5C_Midsag_Composite_byPMA.pdf", fig_s3d, width = 3.42, height = 5.46, dpi = 300)
message("Saved: Figure_S5C_Midsag_Composite_byPMA.{pdf,png}")
```

# Per-group n and pairwise P

```{r summary-table}
plot_df %>% mutate(tp = case_when(xc <= 2.5 ~ "28 wk", xc <= 5 ~ "33 wk", TRUE ~ "35 wk")) %>%
 group_by(tp, GA_Label) %>% summarise(n = n(), median_z = round(median(y),2), .groups = "drop") %>%
 kable(caption = "Per-group n and median midsagittal composite") %>%
 kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

# Session info

```{r session}
sessionInfo()
```
