---
title: "Figure S9 -- GSEA GO Enrichment Heatmap"
subtitle: "Spatial Transcriptomics Pattern-Level Gene Set Enrichment (fgsea, GO Terms)"
author: ""
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
 echo = TRUE,
 message = FALSE,
 warning = FALSE,
 fig.width = 8,
 fig.height = 15,
 dpi = 300
)
knitr::opts_knit$set(root.dir = normalizePath("."))

# --- Location-agnostic dataset finder (RELATIVE paths only) ---
find_dataset <- function(names, subdir = "Supplementary_Datasets", max_up = 6) {
 if (is.character(names) && length(names) == 1) names <- c(names)
 prefix <- "."
 for (i in 0:max_up) {
 for (nm in names) { rel <- file.path(prefix, subdir, nm); if (file.exists(rel)) return(rel) }
 prefix <- file.path(prefix, "..")
 }
 stop("Dataset not found under a 'Supplementary_Datasets/' folder: ", paste(names, collapse = " / "))
}
```

# Overview

**Figure S9** shows a heatmap of Gene Ontology (GO) enrichment results from **fgsea**
(fast gene set enrichment analysis) applied to CoGAPS spatial transcriptomics patterns
from Visium data. Three cerebellar patterns are compared:

| Pattern | Label | Cell Layer |
|---------|-------|-----------|
| Pattern_16 | CMAP_IGL_IMM | Internal Granular Layer (Immature) |
| Pattern_17 | CMAP_IGL | Internal Granular Layer |
| Pattern_27 | CMAP_PCL | Purkinje Cell Layer |

The heatmap displays `log10(-log10(p-value))` for the union of top 50 GO terms per pattern,
with pathways as rows and patterns as columns (vertical layout matching original).

# Libraries

```{r libraries}
library(readxl)
library(dplyr)
library(tidyr)
library(tibble)
library(ComplexHeatmap)
library(circlize)
library(knitr)
library(kableExtra)
```

---

# Source Data & Methods

## Data Provenance

1. **Source**: 10x Visium spatial transcriptomics of postmortem human cerebellar tissue
2. **Pattern decomposition**: CoGAPS (n=30 patterns)
3. **Gene set enrichment**: fgsea with GO gene sets, using gene weights from each pattern
4. **Input file**: `Supplementary_Datasets/Dataset_Visium_GSEA.xlsx` (3 sheets: Pattern_16, Pattern_17, Pattern_27)

Each sheet contains 10,395 GO pathways with columns: `pathway`, `pval`, `padj`, `NES`, `ES`, `size`.

## Data Loading

```{r load-data}
sheets <- c("Pattern_16", "Pattern_17", "Pattern_27")
pattern_labels <- c("Pattern_16" = "CMAP_IGL_IMM",
 "Pattern_17" = "CMAP_IGL",
 "Pattern_27" = "CMAP_PCL")

gsea_file <- find_dataset("Dataset_Visium_GSEA.xlsx")

df <- lapply(sheets, function(s) {
 read_excel(gsea_file, sheet = s) %>%
 select(pathway, pval) %>%
 mutate(Sheet = s)
}) %>%
  bind_rows()
```

---

# Step 1 -- Select Top Pathways

Union of top 50 GO terms per pattern (ranked by p-value), deduplicated.

```{r select-top}
top_pathways <- df %>%
 group_by(Sheet) %>%
 arrange(pval) %>%
 slice_head(n = 50) %>%
 ungroup() %>%
 pull(pathway) %>%
  unique()
```

```{r top-count}
tibble(
 Step = c("Top 50 per pattern (3 patterns)", "Union (deduplicated)"),
 N = c(150, length(top_pathways))
) %>%
 kable(caption = "Pathway selection") %>%
 kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

---

# Step 2 -- Filter to Top Pathways

Filter the full data to include only the union of top-50 pathways per pattern.
Row ordering for the heatmap uses `top_pathways` (preserving first-occurrence order
from grouped `slice_head` + `unique`, matching the original `top_df` variable).

```{r filter-top}
plot_df <- df %>% filter(pathway %in% top_pathways)
```

---

# Step 3 -- Build Heatmap Matrix

Convert to matrix of `-log10(p-value)`, then apply `log10` transformation
(i.e., `log10(-log10(p))`) matching the original figure scale.

```{r build-matrix}
matrix_df <- plot_df %>%
 mutate(neg_log10_p = -log10(pval)) %>%
 select(pathway, Sheet, neg_log10_p) %>%
 pivot_wider(names_from = Sheet, values_from = neg_log10_p) %>%
 column_to_rownames("pathway") %>%
  as.matrix()

# Apply log10 transform (matching original code)
matrix_df <- log10(matrix_df)

# Reorder rows using top_pathways (matching the top_df ordering)
matrix_df <- matrix_df[top_pathways, ]

# Rename columns to biological labels
colnames(matrix_df) <- pattern_labels[colnames(matrix_df)]
```

## Source Data Summary

```{r source-summary}
tibble(
 Pattern = pattern_labels,
 Sheet = names(pattern_labels),
 Top_Pathway = sapply(names(pattern_labels), function(s) {
 df %>% filter(Sheet == s) %>% arrange(pval) %>% slice(1) %>% pull(pathway)
 }),
 Min_pval = sapply(names(pattern_labels), function(s) {
 df %>% filter(Sheet == s) %>% arrange(pval) %>% slice(1) %>% pull(pval) %>%
 sprintf("%.2e", .)
 })
) %>%
 kable(caption = "Most significant GO term per pattern") %>%
 kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

---

# Figure S9 -- GSEA GO Enrichment Heatmap

Same heatmap as the original code, **rotated 90°** to fit the wide layout: the matrix is transposed so the
three patterns are rows (CMAP_PCL / CMAP_IGL / CMAP_IGL_IMM) and the GO terms are columns — columns stay in
the original `top_pathways` order (no reordering). Everything else (colours, legend) is the original format.

```{r fig-S9, fig.width=18, fig.height=13}
mat_t <- t(matrix_df)[c("CMAP_PCL", "CMAP_IGL", "CMAP_IGL_IMM"), ] # rotate; deck row order, columns unchanged

ht <- Heatmap(
 mat_t,
 cluster_rows = FALSE,
 cluster_columns = FALSE,
 row_names_side = "left",
 row_names_gp = gpar(fontsize = 14), # bigger pattern labels
 column_names_gp = gpar(fontsize = 9), # bigger GO-term labels
 width = ncol(mat_t) * unit(3.6, "mm"), # narrower cells
 height = nrow(mat_t) * unit(14, "mm"), # taller cells
 name = "-Log10(P)",
 show_heatmap_legend = FALSE # legend placed manually (bottom-left)
)

# draw the heatmap with extra bottom padding, then place the colour bar low at bottom-left
draw_S9 <- function() {
 draw(ht, padding = unit(c(80, 2, 2, 2), "mm")) # reserve space below the labels
 lgd <- Legend(col_fun = ht@matrix_color_mapping@col_fun, title = "-Log10(P)",
 direction = "horizontal", at = c(-0.5, 0, 0.5, 1, 1.5),
 legend_width = unit(4.5, "cm"), title_position = "lefttop")
 draw(lgd, x = unit(3.5, "cm"), y = unit(6.0, "cm"), just = c("left", "bottom"))
}
draw_S9()
```

```{r pdf-S9, include=FALSE}
pdf("Figure_S9_GSEA_Heatmap.pdf", height = 13, width = 18); draw_S9(); dev.off()
```

---

# Pathway List

```{r pathway-table}
plot_df %>%
 mutate(Sheet = pattern_labels[Sheet]) %>%
 select(Pathway = pathway, Pattern = Sheet, pval) %>%
 pivot_wider(names_from = Pattern, values_from = pval) %>%
 mutate(across(where(is.numeric), ~sprintf("%.2e", .))) %>%
 kable(caption = "P-values for all pathways in the heatmap") %>%
 kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
 scroll_box(height = "500px")
```

---

# Session Info

```{r session-info}
sessionInfo()
```
