---
title: "Figure S6: US Cerebellar Composite Trajectories by GA Group"
subtitle: "Shared-axes scatter plots with linear regression + percent change ANOVA"
author: ""
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
 echo = TRUE, warning = FALSE, message = FALSE,
 fig.width = 14, fig.height = 8, dpi = 300
)
knitr::opts_knit$set(root.dir = normalizePath("."))

# --- Local dataset finder (RELATIVE paths only) ---------------------------
find_dataset <- function(names, subdir = "Supplementary_Datasets", max_up = 6) {
 if (is.character(names) && length(names) == 1) names <- c(names)
 prefix <- "."
 for (i in 0:max_up) {
 for (nm in names) { rel <- file.path(prefix, subdir, nm); if (file.exists(rel)) return(rel) }
 prefix <- file.path(prefix, "..")
 }
 stop("Dataset not found under a 'Supplementary_Datasets/' folder: ", paste(names, collapse = " / "))
}
```

# Data Loading and Composite Calculation

All eight subpanels (EP, VP, LP, All × Coronal, Midsagittal) share the same x-axis (PMA weeks) and a common y-axis within each row, so growth trajectories can be compared visually across GA groups without the eye having to rescale. Limits are applied with `coord_cartesian`, so no data points are dropped and the regression fits and CIs are unchanged.

```{r load_data}
library(readxl)
library(dplyr)
library(ggplot2)
library(patchwork)
library(knitr)
library(kableExtra)

# Color scheme
ga_colors <- c(
 "Extremely Preterm" = "#030403", # ' updated palette (eyedropper)
 "Very Preterm" = "#DE1567",
 "Late Preterm" = "#7BACDC"
)
ga_labels_short <- c("Extremely Preterm" = "EP",
 "Very Preterm" = "VP",
 "Late Preterm" = "LP")

fmt_p <- function(p) {
 if (is.na(p)) return("NA")
 if (p < 0.001) return("< 0.001")
 sprintf("%.3f", p)
}

# --- Load US data ---
data_file <- find_dataset(c("Dataset_Clinical_MRI_US_ND.xlsx", "Dataset_S1_Clinical_MRI_US_ND.xlsx"))
us_raw <- read_excel(data_file, sheet = "Ultrasound")

us_data <- us_raw %>%
 mutate(
 GA_Category = factor(
 case_when(
 GA_weeks < 28 ~ "Extremely Preterm",
 GA_weeks < 32 ~ "Very Preterm",
 TRUE ~ "Late Preterm"
 ),
 levels = c("Extremely Preterm", "Very Preterm", "Late Preterm")
 ),
 Study_ID = as.character(Study_ID),
 across(c(Coronal_Height, Coronal_Transverse, Coronal_Area,
 Midsag_AP, Midsag_Height, Midsag_Area), as.numeric)
 ) %>%
 filter(!is.na(PMA_weeks) & PMA_weeks >= 23 & PMA_weeks <= 45)

# --- IQR outlier detection (k=3) ---
iqr_outlier <- function(x, k = 3) {
 q1 <- quantile(x, 0.25, na.rm = TRUE)
 q3 <- quantile(x, 0.75, na.rm = TRUE)
 iqr <- q3 - q1
 x < (q1 - k * iqr) | x > (q3 + k * iqr)
}

# === CORONAL ===
us_data <- us_data %>%
 mutate(
 IQR_out_Height = iqr_outlier(Coronal_Height, k = 3),
 IQR_out_Trans = iqr_outlier(Coronal_Transverse, k = 3),
 IQR_out_CArea = iqr_outlier(Coronal_Area, k = 3),
 Coronal_Height = ifelse(IQR_out_Height, NA, Coronal_Height),
 Coronal_Transverse = ifelse(IQR_out_Trans, NA, Coronal_Transverse),
 Coronal_Area = ifelse(IQR_out_CArea, NA, Coronal_Area)
 )

cor_cols <- c("Coronal_Height", "Coronal_Transverse", "Coronal_Area")
cor_for_weights <- us_data %>% select(all_of(cor_cols)) %>% filter(complete.cases(.))
cor_mat <- cor(cor_for_weights, use = "pairwise.complete.obs")
mean_r <- sapply(1:3, function(i) mean(abs(cor_mat[i, -i])))
raw_w <- 1 - mean_r
cor_weights <- raw_w / sum(raw_w)
names(cor_weights) <- cor_cols

us_data <- us_data %>%
 mutate(
 Cor_Height_z = as.numeric(scale(Coronal_Height)),
 Cor_Trans_z = as.numeric(scale(Coronal_Transverse)),
 Cor_Area_z = as.numeric(scale(Coronal_Area)),
 US_Coronal_Composite = cor_weights[1] * Cor_Height_z +
 cor_weights[2] * Cor_Trans_z +
 cor_weights[3] * Cor_Area_z
 )

# Filter to valid coronal before midsagittal
us_data <- us_data %>% filter(!is.na(US_Coronal_Composite))

# === MIDSAGITTAL ===
us_data <- us_data %>%
 mutate(
 IQR_out_MAP = iqr_outlier(Midsag_AP, k = 3),
 IQR_out_MHt = iqr_outlier(Midsag_Height, k = 3),
 IQR_out_MAr = iqr_outlier(Midsag_Area, k = 3),
 Midsag_AP = ifelse(IQR_out_MAP, NA, Midsag_AP),
 Midsag_Height = ifelse(IQR_out_MHt, NA, Midsag_Height),
 Midsag_Area = ifelse(IQR_out_MAr, NA, Midsag_Area)
 )

mid_cols <- c("Midsag_AP", "Midsag_Height", "Midsag_Area")
mid_for_weights <- us_data %>% select(all_of(mid_cols)) %>% filter(complete.cases(.))
mid_mat <- cor(mid_for_weights, use = "pairwise.complete.obs")
mid_mean_r <- sapply(1:3, function(i) mean(abs(mid_mat[i, -i])))
mid_raw_w <- 1 - mid_mean_r
mid_weights <- mid_raw_w / sum(mid_raw_w)
names(mid_weights) <- mid_cols

us_data <- us_data %>%
 mutate(
 Mid_AP_z = as.numeric(scale(Midsag_AP)),
 Mid_Ht_z = as.numeric(scale(Midsag_Height)),
 Mid_Area_z = as.numeric(scale(Midsag_Area)),
 US_Midsag_Composite = mid_weights[1] * Mid_AP_z +
 mid_weights[2] * Mid_Ht_z +
 mid_weights[3] * Mid_Area_z
 )

us_data_ms <- us_data %>% filter(!is.na(US_Midsag_Composite))

cat("US data:", nrow(us_data), "scans,", n_distinct(us_data$Study_ID), "patients\n")
cat("By group:", paste(names(table(us_data$GA_Category)), table(us_data$GA_Category), sep = "=", collapse = ", "), "\n")
```

---

# Figure S6: Shared-Axes Scatter Plots Across PMA

Individual US scans plotted against postmenstrual age, split by GA group (EP, VP, LP) and combined ("All"), with linear regression fits and 95% CIs. Every subpanel in a given row uses identical x and y ranges.

```{r compute_shared_limits}
# Shared x-axis: PMA range across the whole US dataset
x_min <- floor(min(us_data$PMA_weeks, na.rm = TRUE))
x_max <- ceiling(max(us_data$PMA_weeks, na.rm = TRUE))

# Shared y-axis for Coronal row
cor_y_min <- floor(min(us_data$US_Coronal_Composite, na.rm = TRUE) * 2) / 2
cor_y_max <- ceiling(max(us_data$US_Coronal_Composite, na.rm = TRUE) * 2) / 2

# Shared y-axis for Midsagittal row
mid_y_min <- floor(min(us_data_ms$US_Midsag_Composite, na.rm = TRUE) * 2) / 2
mid_y_max <- ceiling(max(us_data_ms$US_Midsag_Composite, na.rm = TRUE) * 2) / 2

cat(sprintf("Shared x-axis (PMA): [%g, %g]\n", x_min, x_max))
cat(sprintf("Shared Coronal y-axis: [%g, %g]\n", cor_y_min, cor_y_max))
cat(sprintf("Shared Midsag y-axis: [%g, %g]\n", mid_y_min, mid_y_max))
```

```{r scatter_functions}
# --- Scatter plot function with shared axis limits ---
make_ga_scatter <- function(data, y_var, y_label, ga_group = NULL,
 panel_title = NULL, show_legend = FALSE,
 xlim_shared = NULL, ylim_shared = NULL) {
 if (!is.null(ga_group)) {
 dat <- data %>% filter(GA_Category == ga_group, !is.na(.data[[y_var]]))
 color_val <- ga_colors[ga_group]
 } else {
 dat <- data %>% filter(!is.na(.data[[y_var]]))
 }

 if (nrow(dat) < 3) return(NULL)

 p <- ggplot(dat, aes(x = PMA_weeks, y = .data[[y_var]]))

 if (is.null(ga_group)) {
 # All groups: color by GA
 p <- p +
 geom_point(aes(color = GA_Category), alpha = 1, size = 1.8) +
 geom_smooth(aes(color = GA_Category, fill = GA_Category),
 method = "lm", se = TRUE, alpha = 0.12, linewidth = 1) +
 scale_color_manual(values = ga_colors,
 labels = ga_labels_short, name = NULL) +
 scale_fill_manual(values = ga_colors,
 labels = ga_labels_short, name = NULL)
 } else {
 # Single group: one color
 p <- p +
 geom_point(color = color_val, alpha = 1, size = 1.8) +
 geom_smooth(method = "lm", se = TRUE, color = color_val, fill = color_val,
 alpha = 0.15, linewidth = 1)
 }

 p <- p +
 labs(x = "Post-menstrual age (weeks)", y = y_label, title = panel_title) +
 theme_bw(base_size = 12) +
 theme(
 panel.grid = element_blank(), # no gridlines
 panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.4), # full box
 panel.background = element_rect(fill = "white", colour = NA),
 axis.title = element_text(size = 13),
 axis.text = element_text(size = 11),
 plot.title = element_text(face = "plain", size = 13),
 legend.text = element_text(size = 12),
 legend.position = ifelse(show_legend, "bottom", "none")
 )

 # Enforce shared axis limits without dropping data (coord_cartesian)
 if (!is.null(xlim_shared) || !is.null(ylim_shared)) {
 p <- p + coord_cartesian(xlim = xlim_shared, ylim = ylim_shared)
 }

 return(p)
}

# --- Slope comparison function (for VP-EP, LP-EP annotations) ---
get_slope_comparisons <- function(data, y_var) {
 groups <- levels(data$GA_Category)
  slopes <- list()
 for (g in groups) {
 d <- data %>% filter(GA_Category == g, !is.na(.data[[y_var]]))
 if (nrow(d) > 3) {
 fit <- lm(as.formula(paste(y_var, "~ PMA_weeks")), data = d)
 slopes[[g]] <- list(
 slope = coef(fit)["PMA_weeks"],
 se = summary(fit)$coefficients["PMA_weeks", "Std. Error"],
 n = nrow(d)
 )
 }
 }

  comparisons <- data.frame()
 pairs <- list(c("Very Preterm", "Extremely Preterm"),
 c("Late Preterm", "Extremely Preterm"))
 pair_labels <- c("VP\u2013EP", "LP\u2013EP")

 for (i in seq_along(pairs)) {
 g1 <- pairs[[i]][1]; g2 <- pairs[[i]][2]
 if (g1 %in% names(slopes) & g2 %in% names(slopes)) {
 diff <- slopes[[g1]]$slope - slopes[[g2]]$slope
 se_diff <- sqrt(slopes[[g1]]$se^2 + slopes[[g2]]$se^2)
 ci_lower <- diff - 1.96 * se_diff
 ci_upper <- diff + 1.96 * se_diff
 sig <- ifelse(ci_lower > 0 | ci_upper < 0, "*", "")

 comparisons <- rbind(comparisons, data.frame(
 Label = pair_labels[i],
 Diff = round(diff, 3),
 CI = sprintf("95%%CI (%.1f, %.1f)%s", ci_lower, ci_upper, sig),
 stringsAsFactors = FALSE
 ))
 }
 }
 comparisons
}
```

```{r fig_s4, fig.width=14, fig.height=8}
cor_xlim <- c(x_min, x_max)
cor_ylim <- c(cor_y_min, cor_y_max)
mid_xlim <- c(x_min, x_max)
mid_ylim <- c(mid_y_min, mid_y_max)

# === Row 1: Coronal ===
p_cor_ep <- make_ga_scatter(us_data, "US_Coronal_Composite", "Coronal Composite (Z-score)",
 "Extremely Preterm", "EP",
 xlim_shared = cor_xlim, ylim_shared = cor_ylim)
p_cor_vp <- make_ga_scatter(us_data, "US_Coronal_Composite", "Coronal Composite (Z-score)",
 "Very Preterm", "VP",
 xlim_shared = cor_xlim, ylim_shared = cor_ylim)
p_cor_lp <- make_ga_scatter(us_data, "US_Coronal_Composite", "Coronal Composite (Z-score)",
 "Late Preterm", "LP",
 xlim_shared = cor_xlim, ylim_shared = cor_ylim)
p_cor_all <- make_ga_scatter(us_data, "US_Coronal_Composite", "Coronal Composite (Z-score)",
 NULL, "All",
 xlim_shared = cor_xlim, ylim_shared = cor_ylim)

# Slope comparison annotations on "All" — anchor to shared-axis corner
cor_comps <- get_slope_comparisons(us_data, "US_Coronal_Composite")
if (nrow(cor_comps) > 0) {
 for (j in seq_len(nrow(cor_comps))) {
 p_cor_all <- p_cor_all +
 annotate("text",
 x = cor_xlim[2] - 0.5,
 y = cor_ylim[1] + 0.15 + (j - 1) * 0.35,
 label = paste0(cor_comps$Label[j], ": ", cor_comps$CI[j]),
 hjust = 1, size = 2.8, fontface = "plain")
 }
}

# === Row 2: Midsagittal ===
p_mid_ep <- make_ga_scatter(us_data_ms, "US_Midsag_Composite", "Midsag Composite (Z-score)",
 "Extremely Preterm", "EP",
 xlim_shared = mid_xlim, ylim_shared = mid_ylim)
p_mid_vp <- make_ga_scatter(us_data_ms, "US_Midsag_Composite", "Midsag Composite (Z-score)",
 "Very Preterm", "VP",
 xlim_shared = mid_xlim, ylim_shared = mid_ylim)
p_mid_lp <- make_ga_scatter(us_data_ms, "US_Midsag_Composite", "Midsag Composite (Z-score)",
 "Late Preterm", "LP",
 xlim_shared = mid_xlim, ylim_shared = mid_ylim)
p_mid_all <- make_ga_scatter(us_data_ms, "US_Midsag_Composite", "Midsag Composite (Z-score)",
 NULL, "All",
 xlim_shared = mid_xlim, ylim_shared = mid_ylim)

mid_comps <- get_slope_comparisons(us_data_ms, "US_Midsag_Composite")
if (nrow(mid_comps) > 0) {
 for (j in seq_len(nrow(mid_comps))) {
 p_mid_all <- p_mid_all +
 annotate("text",
 x = mid_xlim[2] - 0.5,
 y = mid_ylim[1] + 0.15 + (j - 1) * 0.35,
 label = paste0(mid_comps$Label[j], ": ", mid_comps$CI[j]),
 hjust = 1, size = 2.8, fontface = "plain")
 }
}

# === Combine 2-row layout ===
row_cor <- (p_cor_ep | p_cor_vp | p_cor_lp | p_cor_all)
row_mid <- (p_mid_ep | p_mid_vp | p_mid_lp | p_mid_all)

fig_s4 <- (row_cor / row_mid) +
 plot_layout(guides = "collect") &
 theme(legend.position = "bottom")

print(fig_s4)

ggsave("Figure_S6_US_Trajectories.pdf",
 fig_s4, width = 14, height = 8, dpi = 300)

# === Save individual panels ===
ggsave("Figure_S6A_Coronal_EP.pdf",
 p_cor_ep, width = 3.5, height = 4, dpi = 300)
ggsave("Figure_S6B_Coronal_VP.pdf",
 p_cor_vp, width = 3.5, height = 4, dpi = 300)
ggsave("Figure_S6C_Coronal_LP.pdf",
 p_cor_lp, width = 3.5, height = 4, dpi = 300)
ggsave("Figure_S6D_Coronal_All.pdf",
 p_cor_all + theme(legend.position = "bottom"),
 width = 3.5, height = 4, dpi = 300)
ggsave("Figure_S6E_Midsag_EP.pdf",
 p_mid_ep, width = 3.5, height = 4, dpi = 300)
ggsave("Figure_S6F_Midsag_VP.pdf",
 p_mid_vp, width = 3.5, height = 4, dpi = 300)
ggsave("Figure_S6G_Midsag_LP.pdf",
 p_mid_lp, width = 3.5, height = 4, dpi = 300)
ggsave("Figure_S6H_Midsag_All.pdf",
 p_mid_all + theme(legend.position = "bottom"),
 width = 3.5, height = 4, dpi = 300)
```

---

# Axis Notes

- All eight panels use the same x-axis range (`r sprintf("[%g, %g]", x_min, x_max)` PMA weeks).
- The four Coronal panels share a y-range of `r sprintf("[%g, %g]", cor_y_min, cor_y_max)`.
- The four Midsagittal panels share a y-range of `r sprintf("[%g, %g]", mid_y_min, mid_y_max)`.
- Limits are applied with `coord_cartesian`, so no data are dropped — the regression fits and CIs are identical to per-panel scaling; only the display ranges change.

---

# Slope Comparison Tables

## Linear Growth Slopes by GA Group

```{r slope_table}
slope_table <- data.frame()

for (comp_info in list(
 list(var = "US_Coronal_Composite", label = "Coronal"),
 list(var = "US_Midsag_Composite", label = "Midsagittal")
)) {
 dat_use <- if (comp_info$var == "US_Midsag_Composite") us_data_ms else us_data
 for (g in levels(us_data$GA_Category)) {
 d <- dat_use %>% filter(GA_Category == g, !is.na(.data[[comp_info$var]]))
 if (nrow(d) > 3) {
 fit <- lm(as.formula(paste(comp_info$var, "~ PMA_weeks")), data = d)
 ci <- confint(fit)["PMA_weeks", ]
 slope_table <- rbind(slope_table, data.frame(
 Composite = comp_info$label,
 GA_Group = ga_labels_short[g],
 N_scans = nrow(d),
 N_patients = n_distinct(d$Study_ID),
 Slope = round(coef(fit)["PMA_weeks"], 4),
 SE = round(summary(fit)$coefficients["PMA_weeks", "Std. Error"], 4),
 CI = sprintf("[%.4f, %.4f]", ci[1], ci[2]),
 p_value = fmt_p(summary(fit)$coefficients["PMA_weeks", "Pr(>|t|)"]),
 stringsAsFactors = FALSE
 ))
 }
 }
}

kable(slope_table, row.names = FALSE,
 col.names = c("Composite", "GA Group", "N Scans", "N Patients",
 "Slope (per wk)", "SE", "95% CI", "p-value"),
 caption = "Linear Growth Slopes by GA Group (OLS)") %>%
 kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Pairwise Slope Differences

```{r slope_diff_table}
all_comps <- rbind(
 cor_comps %>% mutate(Composite = "Coronal"),
 mid_comps %>% mutate(Composite = "Midsagittal")
)

kable(all_comps, row.names = FALSE,
 col.names = c("Comparison", "Slope Diff", "95% CI", "Composite"),
 caption = "Pairwise Slope Differences (B=VP, C=LP, A=EP)") %>%
 kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
 footnote(general = "* indicates 95% CI excludes zero (significant at alpha = 0.05)")
```

---

# Percent Change Between Timepoints with ANOVA

Comparison of change in cerebellar composite z-scores between first and last US scans across GA groups, tested with one-way ANOVA and Tukey HSD post-hoc.

```{r percent_change}
# Identify first and last US scans per patient (patients with >= 2 scans)
patients_multi <- us_data %>%
 group_by(Study_ID) %>%
  filter(n() >= 2) %>%
 ungroup()

first_us <- patients_multi %>%
 group_by(Study_ID) %>%
 slice_min(order_by = PMA_weeks, n = 1, with_ties = FALSE) %>%
 ungroup()

last_us <- patients_multi %>%
 group_by(Study_ID) %>%
 slice_max(order_by = PMA_weeks, n = 1, with_ties = FALSE) %>%
 ungroup()

# Merge first and last
change_data <- first_us %>%
 select(Study_ID, GA_Category,
 PMA_First = PMA_weeks,
 Coronal_First = US_Coronal_Composite,
 Midsag_First = US_Midsag_Composite) %>%
 inner_join(
 last_us %>%
 select(Study_ID,
 PMA_Last = PMA_weeks,
 Coronal_Last = US_Coronal_Composite,
 Midsag_Last = US_Midsag_Composite),
 by = "Study_ID"
 ) %>%
 mutate(
 PMA_Interval = PMA_Last - PMA_First,
 Coronal_Change = Coronal_Last - Coronal_First,
 Midsag_Change = Midsag_Last - Midsag_First,
 Coronal_Rate = Coronal_Change / PMA_Interval,
 Midsag_Rate = Midsag_Change / PMA_Interval
 ) %>%
 filter(PMA_Interval > 0)

cat("Patients with >= 2 scans and PMA interval > 0:", nrow(change_data), "\n")
```

## Summary: Change Between First and Last US

```{r change_summary}
change_summary <- change_data %>%
 group_by(GA_Category) %>%
 summarise(
 N = n(),
 PMA_First = sprintf("%.1f (%.1f)", mean(PMA_First), sd(PMA_First)),
 PMA_Last = sprintf("%.1f (%.1f)", mean(PMA_Last), sd(PMA_Last)),
 Interval_wk = sprintf("%.1f (%.1f)", mean(PMA_Interval), sd(PMA_Interval)),
 Cor_Change = sprintf("%.3f (%.3f)", mean(Coronal_Change, na.rm = TRUE), sd(Coronal_Change, na.rm = TRUE)),
 Cor_Rate = sprintf("%.4f (%.4f)", mean(Coronal_Rate, na.rm = TRUE), sd(Coronal_Rate, na.rm = TRUE)),
 Mid_Change = sprintf("%.3f (%.3f)", mean(Midsag_Change, na.rm = TRUE), sd(Midsag_Change, na.rm = TRUE)),
 Mid_Rate = sprintf("%.4f (%.4f)", mean(Midsag_Rate, na.rm = TRUE), sd(Midsag_Rate, na.rm = TRUE)),
 .groups = "drop"
 )

kable(change_summary,
 col.names = c("GA Group", "N", "First PMA (wk)", "Last PMA (wk)", "Interval (wk)",
 "Coronal \u0394z", "Coronal \u0394z/wk",
 "Midsag \u0394z", "Midsag \u0394z/wk"),
 caption = "Change Between First and Last US Scan by GA Group (Mean (SD))") %>%
 kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## ANOVA: Change in Composite by GA Group

```{r anova_change}
anova_results <- data.frame()

for (measure_info in list(
 list(var = "Coronal_Change", rate_var = "Coronal_Rate", label = "Coronal"),
 list(var = "Midsag_Change", rate_var = "Midsag_Rate", label = "Midsagittal")
)) {
 dat_change <- change_data %>% filter(!is.na(.data[[measure_info$var]]))
 aov_change <- aov(as.formula(paste(measure_info$var, "~ GA_Category")), data = dat_change)
 aov_p <- summary(aov_change)[[1]][["Pr(>F)"]][1]

 tukey_change <- TukeyHSD(aov_change)$GA_Category

 for (pair in rownames(tukey_change)) {
 anova_results <- rbind(anova_results, data.frame(
 Region = measure_info$label,
 Measure = "Absolute change (\u0394z)",
 Comparison = pair,
 Diff = round(tukey_change[pair, "diff"], 4),
 CI = sprintf("[%.4f, %.4f]", tukey_change[pair, "lwr"], tukey_change[pair, "upr"]),
 p_adj = fmt_p(tukey_change[pair, "p adj"]),
 ANOVA_p = fmt_p(aov_p),
 stringsAsFactors = FALSE
 ))
 }

 dat_rate <- change_data %>% filter(!is.na(.data[[measure_info$rate_var]]))
 aov_rate <- aov(as.formula(paste(measure_info$rate_var, "~ GA_Category")), data = dat_rate)
 aov_rate_p <- summary(aov_rate)[[1]][["Pr(>F)"]][1]

 tukey_rate <- TukeyHSD(aov_rate)$GA_Category

 for (pair in rownames(tukey_rate)) {
 anova_results <- rbind(anova_results, data.frame(
 Region = measure_info$label,
 Measure = "Rate of change (\u0394z/wk)",
 Comparison = pair,
 Diff = round(tukey_rate[pair, "diff"], 4),
 CI = sprintf("[%.4f, %.4f]", tukey_rate[pair, "lwr"], tukey_rate[pair, "upr"]),
 p_adj = fmt_p(tukey_rate[pair, "p adj"]),
 ANOVA_p = fmt_p(aov_rate_p),
 stringsAsFactors = FALSE
 ))
 }
}

kable(anova_results, row.names = FALSE,
 col.names = c("Region", "Measure", "Comparison", "Mean Diff", "95% CI",
 "Tukey p (adj)", "ANOVA p"),
 caption = "ANOVA + Tukey HSD: Change Between First and Last US by GA Group") %>%
 kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## ANOVA: Cross-Sectional Comparison at Key Timepoints

To complement the longitudinal change analysis, we also compare composite z-scores across GA groups at specific PMA windows using one-way ANOVA with Tukey HSD post-hoc tests.

```{r anova_timepoint}
timepoint_anova <- data.frame()

for (target_pma in c(28, 33, 35)) {
 window <- 2

 for (comp_info in list(
 list(var = "US_Coronal_Composite", label = "Coronal"),
 list(var = "US_Midsag_Composite", label = "Midsagittal")
 )) {
 dat_tp <- us_data %>%
 filter(PMA_weeks >= (target_pma - window) & PMA_weeks <= (target_pma + window),
 !is.na(.data[[comp_info$var]])) %>%
 group_by(Study_ID) %>%
 mutate(dist = abs(PMA_weeks - target_pma)) %>%
 slice_min(order_by = dist, n = 1, with_ties = FALSE) %>%
 ungroup()

 group_n <- dat_tp %>% group_by(GA_Category) %>% summarise(n = n(), .groups = "drop") %>% filter(n >= 2)
 dat_tp <- dat_tp %>% filter(GA_Category %in% group_n$GA_Category) %>%
 mutate(GA_Category = droplevels(GA_Category))

 if (length(unique(dat_tp$GA_Category)) >= 2) {
 aov_fit <- aov(as.formula(paste(comp_info$var, "~ GA_Category")), data = dat_tp)
 aov_p <- summary(aov_fit)[[1]][["Pr(>F)"]][1]

 if (length(unique(dat_tp$GA_Category)) >= 2) {
 tukey_fit <- TukeyHSD(aov_fit)$GA_Category

 for (pair in rownames(tukey_fit)) {
 timepoint_anova <- rbind(timepoint_anova, data.frame(
 Timepoint = paste0(target_pma, " wk PMA"),
 Region = comp_info$label,
 Comparison = pair,
 N = nrow(dat_tp),
 Diff = round(tukey_fit[pair, "diff"], 3),
 CI = sprintf("[%.3f, %.3f]", tukey_fit[pair, "lwr"], tukey_fit[pair, "upr"]),
 p_adj = fmt_p(tukey_fit[pair, "p adj"]),
 ANOVA_p = fmt_p(aov_p),
 stringsAsFactors = FALSE
 ))
 }
 }
 }
 }
}

kable(timepoint_anova, row.names = FALSE,
 col.names = c("Timepoint", "Region", "Comparison", "N", "Mean Diff",
 "95% CI", "Tukey p (adj)", "ANOVA p"),
 caption = "Cross-Sectional ANOVA + Tukey HSD at Key PMA Timepoints") %>%
 kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# Session Info

```{r session_info}
sessionInfo()
```
