---
title: "Figure 4A & 4E -- Spatial Transcriptomics Heatmaps"
subtitle: "Z-Scored Gene Expression Across Gestational Age: Purkinje (A) and Granule (E) Cell Populations"
author: ""
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
 echo = TRUE,
 message = FALSE,
 warning = FALSE,
 fig.width = 5,
 fig.height = 6,
 dpi = 300
)
knitr::opts_knit$set(root.dir = normalizePath("."))

# --- Local dataset finder (RELATIVE paths only; no user/machine path leaks) ---
find_dataset <- function(names, subdir = "Supplementary_Datasets", max_up = 6) {
 if (is.character(names) && length(names) == 1) names <- c(names)
 prefix <- "."
 for (i in 0:max_up) {
 for (nm in names) { rel <- file.path(prefix, subdir, nm); if (file.exists(rel)) return(rel) }
 prefix <- file.path(prefix, "..")
 }
 stop("Dataset not found under a 'Supplementary_Datasets/' folder: ", paste(names, collapse = " / "))
}
```

# Overview

Panels **A** and **E** display z-scored gene expression heatmaps from **Visium spatial transcriptomics** across 8 postmortem cerebellar samples (GA 26--40 wk). The top 20 differentially expressed genes are shown for each cell population. Genes subsequently validated by RNAScope ISH are highlighted with dashed boxes.

| Panel | Cell Population | Validated Genes |
|-------|----------------|-----------------|
| **A** | Purkinje cells (top 20 genes) | PLP1, KLF9 |
| **E** | Granule cells (top 20 genes) | TUBB4A, GRM4 |

# Libraries

```{r libraries}
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)
```

---

# Source Data & Methods

## Data Source

Z-scored expression values were derived from **10x Visium spatial transcriptomics** of postmortem human cerebellar tissue. Cell-type-specific gene expression was deconvolved, ranked by differential expression, and z-scored across samples.

**Samples**: 8 postmortem cerebellar specimens spanning GA 26, 27, 30, 34, 38, 38, 39, 40 weeks.

**Input file**: `Supplementary_Datasets/Data_S5_InSitu_Validation.xlsx`, sheet `Heatmap_ZScores`.

## Data Loading

```{r load-data}
zscore_df <- read_excel(find_dataset("Data_S5_InSitu_Validation.xlsx"), sheet = "Heatmap_ZScores")
```

## Source Data -- Z-Scored Expression (All Top 20 Genes)

```{r source-data}
zscore_df %>%
 select(Cell_Type, Rank, Gene, GA_26:GA_40) %>%
 kable(digits = 2, caption = "Z-scored expression values for all top 20 genes per cell type") %>%
 kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
 scroll_box(height = "400px")
```

---

# Shared Helpers

```{r helpers}
ga_cols <- c("GA_26", "GA_27", "GA_30", "GA_34", "GA_38", "GA_38b", "GA_39", "GA_40")
ga_labels <- c("26", "27", "30", "34", "38", "38", "39", "40")

# Deck-matched styling: GA axis (numbers + title) placed at TOP; all gene labels
# plain (the dashed box is the sole highlight for RNAScope-validated genes).
build_heatmap <- function(data_top, rnascope_genes) {
 gene_order <- data_top$Gene
 long_df <- data_top %>%
 select(Gene, all_of(ga_cols)) %>%
 pivot_longer(cols = all_of(ga_cols), names_to = "GA_col", values_to = "Zscore") %>%
 mutate(GA_idx = match(GA_col, ga_cols),
 Gene = factor(Gene, levels = rev(gene_order)))

 ggplot(long_df, aes(x = GA_idx, y = Gene, fill = Zscore)) +
 geom_tile(colour = "white", linewidth = 0.3) +
 scale_fill_gradient2(
 low = "#2166AC", mid = "white", high = "#B2182B",
 midpoint = 0, limits = c(-2.5, 2.5),
 name = "Z-score", breaks = c(-2, -1, 0, 1, 2)
 ) +
 scale_x_continuous(breaks = 1:8, labels = ga_labels,
 expand = c(0, 0), position = "top") +
 {
 lapply(rnascope_genes, function(g) {
 ypos <- which(rev(gene_order) == g)
 annotate("rect", xmin = 0.5, xmax = 8.5,
 ymin = ypos - 0.5, ymax = ypos + 0.5,
 fill = NA, colour = "black", linetype = "dashed", linewidth = 0.5)
 })
 } +
 labs(x = "GA", y = NULL) +
 theme_minimal(base_size = 10) +
 theme(
 axis.text.x = element_text(size = 8),
 axis.text.y = element_text(size = 8),
 axis.title.x = element_text(size = 9),
 axis.title.x.top = element_text(size = 9, margin = margin(b = 4)),
 panel.grid = element_blank(),
 legend.position = "right",
 legend.key.height = unit(0.8, "cm"),
 legend.key.width = unit(0.3, "cm"),
 legend.title = element_text(size = 8),
 legend.text = element_text(size = 7),
 plot.margin = margin(5, 5, 5, 5)
 )
}
```

---

# Figure 4A -- Purkinje Cell Heatmap

Top 20 Purkinje cell genes ranked by differential expression. RNAScope-validated genes (**PLP1**, **KLF9**) are highlighted with dashed boxes.

```{r fig-4A, fig.width=5, fig.height=6}
pc_top <- zscore_df %>% filter(Cell_Type == "Purkinje", Rank == "Top")
p4A <- build_heatmap(pc_top, rnascope_genes = c("PLP1", "KLF9"))
print(p4A)
```

```{r pdf-4A, include=FALSE}
ggsave("Figure_4A_Purkinje_Heatmap.pdf", p4A, width = 120, height = 150, units = "mm", dpi = 300)
```

---

# Figure 4E -- Granule Cell Heatmap

Top 20 Granule cell genes ranked by differential expression. RNAScope-validated genes (**TUBB4A**, **GRM4**) are highlighted with dashed boxes.

```{r fig-4E, fig.width=5, fig.height=6}
gc_top <- zscore_df %>% filter(Cell_Type == "Granule", Rank == "Top")
p4E <- build_heatmap(gc_top, rnascope_genes = c("TUBB4A", "GRM4"))
print(p4E)
```

```{r pdf-4E, include=FALSE}
ggsave("Figure_4E_Granule_Heatmap.pdf", p4E, width = 120, height = 150, units = "mm", dpi = 300)
```

---

# Session Info

```{r session-info}
sessionInfo()
```
