---
title: "Figure 1A: Cerebellar Volume vs Gestational Age"
subtitle: "PMA-Adjusted Linear Regression (Volumetric MRI Cohort, N = 120)"
author: ""
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
 echo = TRUE, warning = FALSE, message = FALSE,
 fig.width = 3.2, fig.height = 4.5, dpi = 300
)
knitr::opts_knit$set(root.dir = normalizePath("."))

# --- Local dataset finder (RELATIVE paths only) ---------------------------
# Data lives in Figures_V2_Final/Supplementary_Datasets/ (git-ignored, kept off GitHub). This
# returns a RELATIVE path by walking up to find a "Supplementary_Datasets/" folder, so no
# machine-specific / user path is ever embedded in the code or rendered HTML.
find_dataset <- function(names, subdir = "Supplementary_Datasets", max_up = 6) {
 if (is.character(names) && length(names) == 1) names <- c(names)
 prefix <- "."
 for (i in 0:max_up) {
 for (nm in names) {
 rel <- file.path(prefix, subdir, nm)
 if (file.exists(rel)) return(rel)
 }
 prefix <- file.path(prefix, "..")
 }
 stop("Dataset not found under a 'Supplementary_Datasets/' folder: ", paste(names, collapse = " / "))
}
```

# Data Loading and Exclusion Flow

## Cohort Derivation

The volumetric MRI analytic cohort (N = 120) was derived from the full enrolled sample through the following steps:

```{r exclusion_flow}
library(readxl)
library(dplyr)
library(ggplot2)
library(sandwich)
library(lmtest)
library(car)
library(knitr)
library(kableExtra)

# Exclusion flow table
excl_flow <- data.frame(
 Step = c(
 "MRI scans after quality review",
 "Subjects with automated cerebellar 3D segmentation",
 "Exclusions from segmented cohort",
 "",
 "",
 "",
 "IQR volume outlier screening (k = 3)",
 "**Final volumetric MRI cohort**"
 ),
 Detail = c(
 "",
 "57 subjects lacked automated cerebellar segmentation (Label 6)",
 "ID 469 (GA=33.8, LP): Scan timing -- scanned ~2 days post-birth (PMA 34.1 wk), not TEA",
 "ID 57 (GA=28.0, VP): Scan timing -- scanned 20 wk post-birth (PMA 48.0 wk), outside TEA window",
 "ID 104 (GA=26.4, EP): PMA data-entry error -- MRI_CA=480 wk, computed PMA=56.4 wk",
 "",
 "No additional subjects excluded (all within fences)",
 ""
 ),
 N = c("N = 180", "N = 123", "-1", "-1", "-1", "", "-0", "**N = 120**"),
 stringsAsFactors = FALSE
)

# Display merged exclusion info per step
kable(excl_flow,
 caption = "Exclusion Flow: Volumetric MRI Cohort",
 col.names = c("Step", "Detail", "N")) %>%
 kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
 row_spec(8, bold = TRUE, background = "#f0f7fb")
```

**GA Group Definitions:**

| Group | Label | GA Range |
|-------|-------|----------|
| EP | Extremely Preterm | < 28 weeks |
| VP | Very Preterm | 28 to < 32 weeks |
| LP | Late Preterm | >= 32 weeks |

**Data Source:** `Supplementary_Datasets/Dataset_Clinical_MRI_US_ND.xlsx`, sheet = "MRI". The sheet contains all 123 subjects with automated cerebellar 3D segmentation; 3 subjects are excluded (IDs 57, 104, 469) and marked via the `Included` and `Exclusion_Reason` columns. The code below filters to `Included == "Yes"` to obtain the analytic sample (N = 120).

## Load Data

```{r load_data}
# ============================================================
# LOAD FROM FINAL ANALYSIS DATASET — MRI sheet (volumetric)
# Sheet contains all 123 subjects with segmentation data;
# filter to Included == "Yes" for the analytic sample (N = 120)
# ============================================================
data_file <- find_dataset(c("Dataset_Clinical_MRI_US_ND.xlsx", "Dataset_S1_Clinical_MRI_US_ND.xlsx"))
df_all <- read_excel(data_file, sheet = "MRI")

# Show what was excluded
excluded <- df_all %>% filter(Included == "No")
if (nrow(excluded) > 0) {
 excluded %>%
 select(Study_ID, GA_weeks, GA_Group, MRI_PMA_weeks, Exclusion_Reason) %>%
 kable(caption = "Excluded from Volumetric MRI Analysis",
 col.names = c("Study ID", "GA (wk)", "Group", "PMA (wk)", "Reason")) %>%
 kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
}

# Filter to included subjects only
df <- df_all %>% filter(Included == "Yes")

# Rename for convenience (matching original analysis variable names)
df <- df %>%
 rename(
 GA = GA_weeks,
 MRI_PMA = MRI_PMA_weeks,
 MRI_CA = MRI_CA_weeks,
 CerebVol = Cerebellar_Volume_mm3
 ) %>%
 mutate(
 GA_group = factor(
 case_when(
 GA < 28 ~ "EP",
 GA < 32 ~ "VP",
 TRUE ~ "LP"
 ),
 levels = c("EP", "VP", "LP")
 )
 )

# Center PMA at cohort mean
pma_mean <- mean(df$MRI_PMA, na.rm = TRUE)
df$PMA_c <- df$MRI_PMA - pma_mean

data.frame(
 Parameter = c("Volumetric MRI cohort", "PMA centered at"),
 Value = c(paste("N =", nrow(df)), paste0(round(pma_mean, 1), " weeks"))
) %>%
 kable(col.names = c("", ""), caption = "Analysis Parameters") %>%
 kable_styling(bootstrap_options = c("condensed"), full_width = FALSE)

# Color scheme (' updated palette 2026-08 — eyedropper from PowerPoint)
ga_colors <- c(
 "EP" = "#030403", # near-black
 "VP" = "#DE1567", # crimson/magenta
 "LP" = "#7BACDC" # light blue
)

ga_labels <- c(
 "EP" = "EP",
 "VP" = "VP",
 "LP" = "LP"
)

# Cohort table
df %>%
 group_by(GA_group) %>%
 summarise(
 n = n(),
 GA_mean = sprintf("%.1f (%.1f)", mean(GA), sd(GA)),
 PMA_mean = sprintf("%.1f (%.1f)", mean(MRI_PMA), sd(MRI_PMA)),
 CerebVol_mean = sprintf("%s (%s)",
 formatC(round(mean(CerebVol)), format = "d", big.mark = ","),
 formatC(round(sd(CerebVol)), format = "d", big.mark = ",")),
 .groups = "drop"
 ) %>%
 kable(caption = "Volumetric MRI Cohort",
 col.names = c("GA Group", "n", "GA (wk)", "PMA at MRI (wk)", "Cerebellar Volume (mm³)")) %>%
 kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# Sex as Covariate Check

```{r sex_check}
model_with_sex <- lm(CerebVol ~ PMA_c + GA + Sex, data = df)
sex_summary <- summary(model_with_sex)
sex_p <- sex_summary$coefficients["SexM", "Pr(>|t|)"]
sex_beta <- round(coef(model_with_sex)["SexM"], 1)

data.frame(
 Covariate = "Sex (Male vs Female)",
 Coefficient = paste0(sex_beta, " mm³"),
 `p-value` = round(sex_p, 3),
 Decision = ifelse(sex_p < 0.05, "Include in primary model",
 "Exclude (non-significant)"),
 check.names = FALSE
) %>%
 kable(caption = "Sex as potential covariate") %>%
 kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# Primary Model: CerebVol ~ PMA_c + GA

```{r primary_model}
model_primary <- lm(CerebVol ~ PMA_c + GA, data = df)
model_summary <- summary(model_primary)

# HC3 robust standard errors
robust_se <- sqrt(diag(vcovHC(model_primary, type = "HC3")))
robust_t <- coef(model_primary) / robust_se
robust_p <- 2 * pt(abs(robust_t), df = model_primary$df.residual, lower.tail = FALSE)

# Coefficient table
coef_tbl <- as.data.frame(coef(model_summary))
results_df <- data.frame(
 Variable = c("Intercept", "PMA at MRI (centered)", "GA at birth"),
 Coefficient = round(coef(model_primary), 0),
 SE = round(coef_tbl$`Std. Error`, 0),
 `p-value` = ifelse(coef_tbl$`Pr(>|t|)` < 0.001, "< 0.001",
 round(coef_tbl$`Pr(>|t|)`, 3)),
 `p (HC3)` = ifelse(robust_p < 0.001, "< 0.001", round(robust_p, 3)),
 check.names = FALSE
)

kable(results_df,
 caption = paste0("Primary Model Results (R² = ",
 round(model_summary$r.squared, 3),
 ", Adj R² = ",
 round(model_summary$adj.r.squared, 3), ")")) %>%
 kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Effect Sizes

```{r effect_sizes}
# Partial r for GA (after controlling for PMA)
t_GA <- coef(model_summary)["GA", "t value"]
partial_r_GA <- t_GA / sqrt(t_GA^2 + model_primary$df.residual)

# Cohen's f² for GA
model_reduced <- lm(CerebVol ~ PMA_c, data = df)
r2_full <- model_summary$r.squared
r2_reduced <- summary(model_reduced)$r.squared
f2_GA <- (r2_full - r2_reduced) / (1 - r2_full)

es_tbl <- data.frame(
 Metric = c("Partial r (GA | PMA)",
 "Cohen's f² (GA)",
 "Effect interpretation",
 "R² (full model)",
 "R² (PMA only)",
 "Incremental R² from GA"),
 Value = c(round(partial_r_GA, 2),
 round(f2_GA, 3),
 ifelse(f2_GA >= 0.35, "Large",
 ifelse(f2_GA >= 0.15, "Medium",
 ifelse(f2_GA >= 0.02, "Small", "Negligible"))),
 round(r2_full, 3),
 round(r2_reduced, 3),
 round(r2_full - r2_reduced, 3))
)

kable(es_tbl, caption = "Effect Sizes", col.names = c("Metric", "Value")) %>%
 kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Model Diagnostics

```{r diagnostics}
shapiro_test <- shapiro.test(residuals(model_primary))
bp_test <- bptest(model_primary)
vif_vals <- vif(model_primary)

diag_df <- data.frame(
 Test = c("Shapiro-Wilk (normality)", "Breusch-Pagan (homoscedasticity)",
 "VIF: PMA_c", "VIF: GA"),
 Statistic = c(round(shapiro_test$statistic, 3), round(bp_test$statistic, 2),
 round(vif_vals["PMA_c"], 2), round(vif_vals["GA"], 2)),
 `p-value` = c(round(shapiro_test$p.value, 3), round(bp_test$p.value, 3), "—", "—"),
 check.names = FALSE
)

kable(diag_df, caption = "Model Diagnostics") %>%
 kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Sensitivity Analysis (Cook's Distance)

```{r sensitivity}
cooks_d <- cooks.distance(model_primary)
threshold <- 4 / nrow(df)
n_influential <- sum(cooks_d > threshold)

df_sens <- df[cooks_d <= threshold, ]
model_sensitivity <- lm(CerebVol ~ PMA_c + GA, data = df_sens)
sens_summary <- summary(model_sensitivity)

sens_df <- data.frame(
 Model = c("Primary", "Sensitivity (excl. influential)"),
 n = c(nrow(df), nrow(df_sens)),
 `GA Coefficient` = c(round(coef(model_primary)["GA"], 0),
 round(coef(model_sensitivity)["GA"], 0)),
 `GA p-value` = c(
 ifelse(coef(model_summary)["GA", "Pr(>|t|)"] < 0.001, "< 0.001",
 round(coef(model_summary)["GA", "Pr(>|t|)"], 3)),
 ifelse(sens_summary$coefficients["GA", "Pr(>|t|)"] < 0.001, "< 0.001",
 round(sens_summary$coefficients["GA", "Pr(>|t|)"], 3))),
 R_squared = c(round(model_summary$r.squared, 3), round(sens_summary$r.squared, 3)),
 check.names = FALSE
)

kable(sens_df, caption = paste0("Sensitivity Analysis (Cook's D threshold = ",
 round(threshold, 4), "; ",
 n_influential, " influential observations)")) %>%
 kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# Figure 1B

```{r figure_1b, fig.width=3.2, fig.height=4.5}
# ============================================================
# FIGURE 1B: Cerebellar Volume vs GA at Birth
# Model: CerebVol ~ PMA_c + GA
# Regression line: predicted at mean PMA (PMA_c = 0)
# ============================================================

beta_ga <- coef(model_primary)["GA"]

# Predicted line at mean PMA
ga_seq <- seq(min(df$GA), max(df$GA), length.out = 100)
pred_df <- data.frame(GA = ga_seq, PMA_c = 0)
pred_ci <- predict(model_primary, newdata = pred_df, interval = "confidence")
pred_df$pred <- pred_ci[, "fit"]
pred_df$lwr <- pred_ci[, "lwr"]
pred_df$upr <- pred_ci[, "upr"]

# Annotation values
partial_r <- round(partial_r_GA, 2)
p_val <- coef(model_summary)["GA", "Pr(>|t|)"]
p_display <- ifelse(p_val < 0.001, "< 0.001", round(p_val, 3))
beta_ga_val <- round(beta_ga, 0)

# Build annotation: two parsed expressions stacked vertically
# Position in lower-right of plot area
# Stat box: right-anchored to the panel edge so it fits the narrow panel
box_xr <- max(df$GA) + 0.3
box_xl <- box_xr - 7.0 # snug box width (roomier for larger text)
box_cx <- (box_xl + box_xr) / 2 # centre for centred text
box_y_base <- min(df$CerebVol) - diff(range(df$CerebVol)) * 0.03
y_step <- diff(range(df$CerebVol)) * 0.04

stat_exprs <- c(
 paste0("beta == ", beta_ga_val, " ~ mm^3 ~ wk^-1"),
 paste0("italic(r) == ", partial_r, "*',' ~~ italic(P) == ", p_display)
)

fig_1b <- ggplot() +
 geom_ribbon(data = pred_df, aes(x = GA, ymin = lwr, ymax = upr),
 fill = "gray80", alpha = 0.25) +
 geom_line(data = pred_df, aes(x = GA, y = pred),
 colour = "black", linewidth = 0.4) +
 geom_point(data = df, aes(x = GA, y = CerebVol, colour = GA_group),
 size = 1.2, alpha = 1) +
 # Stat annotation (no box) — plain text, matching the deck
 annotate("text", x = box_cx, y = box_y_base + y_step * 1.5,
 label = stat_exprs[1], parse = TRUE,
 size = 2.4, fontface = "bold", hjust = 0.5) +
 annotate("text", x = box_cx, y = box_y_base + y_step * 0.2,
 label = stat_exprs[2], parse = TRUE,
 size = 2.4, fontface = "bold", hjust = 0.5) +
 scale_colour_manual(values = ga_colors, labels = ga_labels, name = NULL) +
 scale_x_continuous(breaks = c(22, 26, 30, 34)) +
 scale_y_continuous(labels = scales::comma, breaks = seq(10000, 25000, 5000),
 expand = expansion(mult = c(0.05, 0.02))) +
 labs(
 x = "Gestational age at birth (weeks)",
 y = expression("Cerebellar volume (mm"^3*")")
 ) +
 theme_minimal(base_size = 10) +
 theme(
 legend.position = "bottom",
 legend.margin = margin(t = -2, b = 0),
 legend.key.size = unit(0.3, "cm"),
 legend.text = element_text(size = 8),
 axis.title.x = element_text(size = 10),
 axis.title.y = element_text(size = 10),
 axis.text = element_text(size = 10),
 axis.line = element_blank(),
 panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5),
 panel.grid = element_blank(),
 plot.margin = margin(t = 3, r = 5, b = 3, l = 3)
 )

fig_1b

# Save PDF
ggsave("Figure_1A_CerebVol_vs_GA.pdf",
 fig_1b, width = 2.75, height = 4, dpi = 300)
message("Saved: Figure_1/Figure_1A_CerebVol_vs_GA.pdf")
```

## Statistical Summary Table

```{r stats_table}
fig_stats <- data.frame(
 Metric = c("GA coefficient",
 "Partial r (GA | PMA)",
 "p-value (OLS)",
 "p-value (HC3 robust)",
 "Cohen's f² (GA)",
 "Model R²",
 "Sample size"),
 Value = c(
 paste0(beta_ga_val, " mm³/week"),
 round(partial_r_GA, 2),
 ifelse(p_val < 0.001, "< 0.001", round(p_val, 3)),
 ifelse(robust_p["GA"] < 0.001, "< 0.001", round(robust_p["GA"], 3)),
 round(f2_GA, 3),
 round(r2_full, 3),
 nrow(df)
 )
)

kable(fig_stats, caption = "Figure 1B — Statistical Summary",
 col.names = c("Metric", "Value")) %>%
 kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# Session Info

```{r session_info}
sessionInfo()
```
