---
title: "Figure 4 -- Cerebellar Layer Analysis"
subtitle: "Preterm vs Term: Gestational Age and Chronological Age"
author: ""
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
 echo = TRUE,
 warning = FALSE,
 message = FALSE,
 fig.align = "center",
 dpi = 300,
 results = 'asis'
)
knitr::opts_knit$set(root.dir = normalizePath("."))

# --- Local dataset finder (RELATIVE paths only) ---------------------------
find_dataset <- function(names, subdir = "Supplementary_Datasets", max_up = 6) {
 if (is.character(names) && length(names) == 1) names <- c(names)
 prefix <- "."
 for (i in 0:max_up) {
 for (nm in names) { rel <- file.path(prefix, subdir, nm); if (file.exists(rel)) return(rel) }
 prefix <- file.path(prefix, "..")
 }
 stop("Dataset not found under a 'Supplementary_Datasets/' folder: ", paste(names, collapse = " / "))
}
```

# Overview

**Figure 4 (Panels B--E)** shows cerebellar layer measurements across gestational and chronological age for preterm vs term infants. Each panel pair consists of:

- **Left**: All subjects vs gestational age (single trend line)
- **Right**: Preterm vs term separated by chronological age (separate trend lines with SE ribbons)

| Panel | Layer | Measurement |
|-------|-------|-------------|
| **B** | EGL | External Granular Layer surface area (mm²/mm ML) |
| **C** | ML | Molecular Layer surface area (mm²/mm ML) |
| **D** | IGL | Internal Granular Layer surface area (mm²/mm ML) |
| **E** | PC | Purkinje Cell density (PCs/mm ML) |

# Libraries

```{r libraries, include=FALSE}
library(tidyverse)
library(readxl)
library(gridExtra)
library(grid)
library(ggpubr)
library(cowplot)
library(ggtext)
library(knitr)
library(kableExtra)
```

---

# Source Data & Methods

## Data Provenance

1. **Source**: Histological sections from preterm (<37 wks GA) and term (≥37 wks GA) cerebellar tissue
2. **Quantification**: Deep learning-based layer segmentation from H&E-stained WSIs
3. **Normalization**: Layer areas normalized to molecular layer length; PC density per mm ML
4. **Input file**: `Datasets/Dataset_Cerebellar_Layers.xlsx`, sheet = "Cerebellar_Layers"

## Data Loading

```{r load-data, include=FALSE}
data <- read_excel(find_dataset(c("Dataset_Cerebellar_Layers.xlsx")), sheet = "Cerebellar_Layers")

data <- data %>%
 mutate(Group = factor(Group, levels = c("Preterm", "Term")))

n_total <- n_distinct(data$Subject_ID)
n_preterm <- n_distinct(data$Subject_ID[data$Group == "Preterm"])
n_term <- n_distinct(data$Subject_ID[data$Group == "Term"])
```

```{r data-summary-table}
data %>%
 group_by(Group) %>%
 summarise(
 `N (rows)` = n(),
 `N (subjects)` = n_distinct(Subject_ID),
 `GA (mean ± SD)` = sprintf("%.1f ± %.1f", mean(GA, na.rm = TRUE), sd(GA, na.rm = TRUE)),
 `CA (mean ± SD)` = sprintf("%.1f ± %.1f", mean(Chronological_age, na.rm = TRUE), sd(Chronological_age, na.rm = TRUE)),
 .groups = "drop"
 ) %>%
 kable(caption = "Sample Characteristics by Group", align = c("l", "c", "c", "c", "c")) %>%
 kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

---

# Statistical Functions

```{r statistical-functions, include=FALSE}
# Function to test slope difference between two groups
test_slope_difference <- function(x1, y1, x2, y2) {
 n1 <- sum(!is.na(x1) & !is.na(y1))
 n2 <- sum(!is.na(x2) & !is.na(y2))

 if (n1 < 3 || n2 < 3) {
 return(list(
 slope1 = NA, se1 = NA, r2_1 = NA,
 slope2 = NA, se2 = NA, r2_2 = NA,
 slope_diff = NA, se_diff = NA,
 t_stat = NA, df = NA, p_value = NA
 ))
 }

 valid1 <- !is.na(x1) & !is.na(y1)
 valid2 <- !is.na(x2) & !is.na(y2)
 x1 <- x1[valid1]; y1 <- y1[valid1]
 x2 <- x2[valid2]; y2 <- y2[valid2]

 model1 <- lm(y1 ~ x1)
 slope1 <- coef(model1)[2]
 se1 <- summary(model1)$coefficients[2, 2]
 r2_1 <- summary(model1)$r.squared

 model2 <- lm(y2 ~ x2)
 slope2 <- coef(model2)[2]
 se2 <- summary(model2)$coefficients[2, 2]
 r2_2 <- summary(model2)$r.squared

 slope_diff <- slope1 - slope2
 se_diff <- sqrt(se1^2 + se2^2)
 t_stat <- slope_diff / se_diff
 df <- n1 + n2 - 4
 p_value <- 2 * pt(abs(t_stat), df, lower.tail = FALSE)

 list(
 slope1 = slope1, se1 = se1, r2_1 = r2_1,
 slope2 = slope2, se2 = se2, r2_2 = r2_2,
 slope_diff = slope_diff, se_diff = se_diff,
 t_stat = t_stat, df = df, p_value = p_value
 )
}

# Function for mean difference test
test_mean_difference <- function(y1, y2) {
 y1 <- y1[!is.na(y1)]
 y2 <- y2[!is.na(y2)]
 if (length(y1) < 2 || length(y2) < 2) {
 return(list(p_value = NA, t_stat = NA, mean1 = NA, mean2 = NA))
 }
 test_result <- t.test(y1, y2)
 list(
 p_value = test_result$p.value,
 t_stat = test_result$statistic,
 mean1 = mean(y1),
 mean2 = mean(y2),
 sd1 = sd(y1),
 sd2 = sd(y2)
 )
}

# Format p-value for display (uppercase P, AAAS style)
format_pvalue <- function(p, digits = 3) {
 sapply(p, function(pval) {
 if (is.na(pval)) return("NA")
 if (pval < 0.001) return("P < 0.001***")
 if (pval < 0.01) return(paste0("P = ", format(round(pval, digits), nsmall = digits), "**"))
 if (pval < 0.05) return(paste0("P = ", format(round(pval, digits), nsmall = digits), "*"))
 paste0("P = ", format(round(pval, digits), nsmall = digits))
 })
}
```

---

# Figure 4 -- Cerebellar Layer Analysis (Panels B--E)

```{r create-figure, fig.width=12, fig.height=8.5, echo=FALSE}
# ============================================================
# COLORS
# ============================================================
color_preterm <- "#030403" # black preterm points (' updated palette 2026-08)
color_term <- "#DE1567" # crimson term points
line_preterm <- "#030403" # black preterm trend line
line_term <- "#DE1567" # crimson term trend line

# Line and point settings
line_thickness <- 0.8
point_size <- 2.0
point_alpha <- 1

# ============================================================
# PANEL CREATION FUNCTION
# ============================================================
create_panel_pair <- function(data, measurement_col, y_lab) {

 df <- data %>%
 filter(!is.na(.data[[measurement_col]]), !is.na(GA), !is.na(Chronological_age))

 if (nrow(df) < 10) return(NULL)

 preterm_df <- df %>% filter(Group == "Preterm")
 term_df <- df %>% filter(Group == "Term")

 slope_test <- test_slope_difference(
 preterm_df$Chronological_age, preterm_df[[measurement_col]],
 term_df$Chronological_age, term_df[[measurement_col]]
 )

 # Individual slope coefficients, SE, CI, and p-values (CA)
 model_preterm_ca <- lm(preterm_df[[measurement_col]] ~ preterm_df$Chronological_age)
 model_term_ca <- lm(term_df[[measurement_col]] ~ term_df$Chronological_age)
 s_preterm <- summary(model_preterm_ca)$coefficients
 s_term <- summary(model_term_ca)$coefficients
 slope_preterm_ca <- s_preterm[2, 1]
 se_preterm_ca <- s_preterm[2, 2]
 p_slope_preterm_ca <- s_preterm[2, 4]
 ci_lo_preterm <- slope_preterm_ca - 1.96 * se_preterm_ca
 ci_hi_preterm <- slope_preterm_ca + 1.96 * se_preterm_ca
 slope_term_ca <- s_term[2, 1]
 se_term_ca <- s_term[2, 2]
 p_slope_term_ca <- s_term[2, 4]
 ci_lo_term <- slope_term_ca - 1.96 * se_term_ca
 ci_hi_term <- slope_term_ca + 1.96 * se_term_ca

 # Overall slope p-value (GA, all subjects)
 model_all_ga <- lm(df[[measurement_col]] ~ df$GA)
 p_slope_all_ga <- summary(model_all_ga)$coefficients[2, 4]

 y_min <- min(df[[measurement_col]], na.rm = TRUE)
 y_max <- max(df[[measurement_col]], na.rm = TRUE)
 y_range <- y_max - y_min
 y_limits <- c(y_min - y_range * 0.08, y_max + y_range * 0.22)

 # P-value formatters (uppercase P, AAAS style)
 fmt_p <- function(p) {
 if (is.na(p)) return("italic(P) == 'NA'")
 if (p < 0.001) return("italic(P) < \"0.001\"")
 paste0('italic(P) == "', sprintf("%.3f", p), '"')
 }

 fmt_p_beta <- function(p) {
 if (is.na(p)) return("italic(P) == 'NA'")
 if (p < 0.0001) return("italic(P) < \"0.0001\"")
 if (p < 0.001) return("italic(P) < \"0.001\"")
 paste0('italic(P) == "', sprintf("%.3f", p), '"')
 }

 # ========================================
 # LEFT PANEL: GA vs Measurement
 # ========================================
 # Fit one OLS across all subjects, then split the line display at GA = 37
 # so the preterm portion is red and the term portion is blue. The CI ribbon
 # remains a single ribbon derived from the overall fit.
 fit_df <- data.frame(GA = df$GA, y = df[[measurement_col]])
 fit_all <- lm(y ~ GA, data = fit_df)
 ga_seq <- seq(min(df$GA), max(df$GA), length.out = 200)
 pred_all <- predict(fit_all, newdata = data.frame(GA = ga_seq),
 interval = "confidence")
 pred_df <- data.frame(GA = ga_seq,
 fit = pred_all[, "fit"],
 lwr = pred_all[, "lwr"],
 upr = pred_all[, "upr"],
 Segment = ifelse(ga_seq < 37, "Preterm", "Term"))

 p_ga <- ggplot(df, aes(x = GA, y = .data[[measurement_col]], color = Group)) +
 geom_point(size = point_size, alpha = point_alpha) +
 geom_ribbon(data = pred_df, inherit.aes = FALSE,
 aes(x = GA, ymin = lwr, ymax = upr),
 fill = "gray60", alpha = 0.15) +
 geom_line(data = pred_df, inherit.aes = FALSE,
 aes(x = GA, y = fit),
 color = "gray30", linewidth = line_thickness) +
 geom_vline(xintercept = 37, linetype = "dashed", color = "gray70", linewidth = 0.5) +
 scale_color_manual(values = c("Preterm" = color_preterm, "Term" = color_term)) +
 scale_x_continuous(limits = c(20, 45), breaks = seq(20, 45, 5)) +
 scale_y_continuous(limits = y_limits) +
 labs(x = "Gestational age at birth (weeks)", y = y_lab) +
 theme_bw(base_size = 10) +
 theme(
 panel.grid.major = element_blank(),
 panel.grid.minor = element_blank(),
 panel.border = element_rect(color = "gray30", linewidth = 0.5),
 axis.title.x = element_text(size = 9, color = "gray30"),
 axis.title.y = element_text(size = 9, color = "gray30"),
 axis.text = element_text(size = 8, color = "gray30"),
 legend.position = "none",
 plot.margin = margin(8, 4, 4, 4)
 ) +
 annotate("text", parse = TRUE, x = 44, y = y_max + y_range * 0.16,
 label = paste0('"Slope"~~', fmt_p(p_slope_all_ga)),
 hjust = 1, size = 2.7, color = "gray30")

 # ========================================
 # RIGHT PANEL: CA vs Measurement
 # ========================================
 p_val <- slope_test$p_value
 p_label <- fmt_p(p_val)

 # --- Text placement: tight top-right corner ---
 txt_x <- 60; txt_hjust <- 1
 txt_y1 <- y_max + y_range * 0.22
 txt_y2 <- y_max + y_range * 0.15
 txt_y3 <- y_max + y_range * 0.08

 p_ca <- ggplot(df, aes(x = Chronological_age, y = .data[[measurement_col]])) +
 geom_point(aes(color = Group), size = point_size, alpha = point_alpha) +
 # Separate trend lines per group with SE ribbons
 geom_smooth(data = preterm_df, aes(x = Chronological_age, y = .data[[measurement_col]]),
 method = "lm", se = TRUE, color = line_preterm,
 fill = line_preterm, alpha = 0.15,
 linetype = "solid", linewidth = line_thickness) +
 geom_smooth(data = term_df, aes(x = Chronological_age, y = .data[[measurement_col]]),
 method = "lm", se = TRUE, color = line_term,
 fill = line_term, alpha = 0.15,
 linetype = "solid", linewidth = line_thickness) +
 scale_color_manual(values = c("Preterm" = color_preterm, "Term" = color_term)) +
 scale_x_continuous(limits = c(0, 60), breaks = seq(0, 60, 10)) +
 scale_y_continuous(limits = y_limits) +
 labs(x = "Chronological age (weeks)", y = y_lab) +
 theme_bw(base_size = 10) +
 theme(
 panel.grid.major = element_blank(),
 panel.grid.minor = element_blank(),
 panel.border = element_rect(color = "gray30", linewidth = 0.5),
 axis.title.x = element_text(size = 9, color = "gray30"),
 axis.title.y = element_text(size = 9, color = "gray30"),
 axis.text = element_text(size = 8, color = "gray30"),
 legend.position = "none",
 plot.margin = margin(8, 4, 4, 4)
 ) +
 annotate("text", parse = TRUE, x = txt_x, y = txt_y1,
 label = paste0('"Slope"~~', p_label),
 hjust = txt_hjust, size = 2.7, color = "gray30") +
 annotate("text", parse = TRUE, x = txt_x, y = txt_y2,
 label = paste0("paste('Preterm: ', italic(beta), ' = ",
 sprintf("%.2f", slope_preterm_ca),
 " [", sprintf("%.2f", ci_lo_preterm), "\u2013", sprintf("%.2f", ci_hi_preterm), "], ', ",
 fmt_p_beta(p_slope_preterm_ca), ")"),
 hjust = txt_hjust, size = 2.7, color = line_preterm) +
 annotate("text", parse = TRUE, x = txt_x, y = txt_y3,
 label = paste0("paste('Term: ', italic(beta), ' = ",
 sprintf("%.2f", slope_term_ca),
 " [", sprintf("%.2f", ci_lo_term), "\u2013", sprintf("%.2f", ci_hi_term), "], ', ",
 fmt_p_beta(p_slope_term_ca), ")"),
 hjust = txt_hjust, size = 2.7, color = line_term)

 list(p_ga = p_ga, p_ca = p_ca, stats = slope_test)
}

# ============================================================
# COLORED HEADER STRIPS
# ============================================================
create_header <- function(title, bg_color, text_color = "white") {
 ggplot() + theme_void() +
 annotate("rect", xmin = 0, xmax = 1, ymin = 0, ymax = 1, fill = bg_color, color = NA) +
 annotate("text", x = 0.5, y = 0.5, label = title, color = text_color, size = 4.5, fontface = "plain") +
 coord_cartesian(xlim = c(0, 1), ylim = c(0, 1), expand = FALSE)
}

create_label <- function(label_text) {
 ggplot() + theme_void() +
 annotate("text", x = 0.5, y = 0.5, label = label_text, size = 5, fontface = "bold", color = "black") +
 coord_cartesian(xlim = c(0, 1), ylim = c(0, 1), expand = FALSE)
}

# EGL and IGL header bars in neutral gray (#CBCED6) per request; ML and PC keep
# their original colors. Gray bars use black text for legibility.
header_egl <- create_header("EGL", "#CBCED6", text_color = "black")
header_ml <- create_header("ML", "#CBCED6", text_color = "black")
header_igl <- create_header("IGL", "#CBCED6", text_color = "black")
header_pc <- create_header("PC", "#CBCED6", text_color = "black")

label_b <- create_label("")
label_c <- create_label("")
label_d <- create_label("")
label_e <- create_label("")

# ============================================================
# CREATE ALL PANELS
# ============================================================
egl_panels <- create_panel_pair(data, "EGL_normalized",
 expression(EGL~surface~area~(mm^2/mu*m~ML)))
ml_panels <- create_panel_pair(data, "IGL_normalized", # deck ML = dataset IGL_normalized (flat)
 expression(ML~surface~area~(mm^2/mu*m~ML)))
igl_panels <- create_panel_pair(data, "ML_normalized", # deck IGL = dataset ML_normalized (strong growth)
 expression(IGL~surface~area~(mm^2/mu*m~ML)))
pc_panels <- create_panel_pair(data, "PC_count",
 expression(PC~Density~(PCs/mm~ML)))

# ============================================================
# COMMON LEGEND
# ============================================================
dummy_legend <- ggplot(data.frame(x = 1:2, y = 1:2,
 Group = factor(c("Preterm", "Term"),
 levels = c("Preterm", "Term"))),
 aes(x, y, color = Group)) +
 geom_point(size = 3, shape = 16) +
 scale_color_manual(values = c("Preterm" = color_preterm, "Term" = color_term)) +
 guides(color = guide_legend(override.aes = list(shape = 16, size = 3))) +
 theme(legend.position = "bottom",
 legend.title = element_blank(),
 legend.text = element_text(size = 9),
 legend.key.size = unit(0.4, "cm"))

shared_legend <- cowplot::get_legend(dummy_legend)

# ============================================================
# ARRANGE FINAL FIGURE
# ============================================================
create_full_panel <- function(label, header, p_ga, p_ca) {
 header_row <- ggarrange(label, header, ncol = 2, widths = c(0.04, 0.96))
 plots_row <- ggarrange(p_ga, p_ca, ncol = 2, widths = c(1, 1))
 ggarrange(header_row, plots_row, nrow = 2, heights = c(0.055, 1))
}

panel_b <- create_full_panel(label_b, header_egl, egl_panels$p_ga, egl_panels$p_ca)
panel_c <- create_full_panel(label_c, header_ml, ml_panels$p_ga, ml_panels$p_ca)
panel_d <- create_full_panel(label_d, header_igl, igl_panels$p_ga, igl_panels$p_ca)
panel_e <- create_full_panel(label_e, header_pc, pc_panels$p_ga, pc_panels$p_ca)

panels_grid <- ggarrange(panel_b, panel_d, panel_c, panel_e,
 ncol = 2, nrow = 2)

final_figure <- ggarrange(panels_grid, shared_legend,
 nrow = 2, heights = c(1, 0.04))

print(final_figure)
```

```{r save-figure, include=FALSE}
ggsave("Figure_3_Cerebellar_Layers.pdf",
 final_figure, width = 12, height = 8.5, units = "in", dpi = 300)
```

---

# Statistical Results Summary

```{r calculate-all-stats, include=FALSE}
measurements <- list(
 EGL = list(col = "EGL_normalized", label = "EGL (External Granular Layer)"),
 ML = list(col = "ML_normalized", label = "ML (Molecular Layer)"),
 IGL = list(col = "IGL_normalized", label = "IGL (Internal Granular Layer)"),
 PC = list(col = "PC_count", label = "PC (Purkinje Cell Density)")
)

all_results <- lapply(names(measurements), function(name) {
 m <- measurements[[name]]
 df <- data %>% filter(!is.na(.data[[m$col]]), !is.na(GA), !is.na(Chronological_age))
 preterm_df <- df %>% filter(Group == "Preterm")
 term_df <- df %>% filter(Group == "Term")

 mean_test <- test_mean_difference(preterm_df[[m$col]], term_df[[m$col]])
 slope_test <- test_slope_difference(
 preterm_df$Chronological_age, preterm_df[[m$col]],
 term_df$Chronological_age, term_df[[m$col]]
 )

 tibble(
 Layer = name,
 `N Total` = nrow(df),
 `N Preterm` = nrow(preterm_df),
 `N Term` = nrow(term_df),
 `Preterm Mean` = mean_test$mean1,
 `Preterm SD` = mean_test$sd1,
 `Term Mean` = mean_test$mean2,
 `Term SD` = mean_test$sd2,
 `Mean Diff p` = mean_test$p_value,
 `Preterm Slope` = slope_test$slope1,
 `Preterm SE` = slope_test$se1,
 `Term Slope` = slope_test$slope2,
 `Term SE` = slope_test$se2,
 `Slope Diff p` = slope_test$p_value
 )
})

results_table <- bind_rows(all_results)
```

## Descriptive Statistics

```{r desc-stats-table}
results_table %>%
 transmute(
 Layer,
 `N Total`,
 `N Preterm`,
 `N Term`,
 `Preterm (Mean ± SD)` = sprintf("%.2f ± %.2f", `Preterm Mean`, `Preterm SD`),
 `Term (Mean ± SD)` = sprintf("%.2f ± %.2f", `Term Mean`, `Term SD`)
 ) %>%
 kable(caption = "Descriptive Statistics by Cerebellar Layer and Group", align = c("l", "c", "c", "c", "c", "c")) %>%
 kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
 add_header_above(c(" " = 1, "Sample Size" = 3, "Measurements" = 2))
```

## Statistical Tests

```{r stat-test-table}
results_table %>%
 transmute(
 Layer,
 `Mean Diff (P)` = format_pvalue(`Mean Diff p`),
 `Preterm Slope (SE)` = sprintf("%.4f (%.4f)", `Preterm Slope`, `Preterm SE`),
 `Term Slope (SE)` = sprintf("%.4f (%.4f)", `Term Slope`, `Term SE`),
 `Slope Diff (P)` = format_pvalue(`Slope Diff p`)
 ) %>%
 kable(caption = "Statistical Test Results: Mean Difference and Slope Difference Tests", align = c("l", "c", "c", "c", "c")) %>%
 kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
 add_header_above(c(" " = 1, "t-test" = 1, "Regression Slopes (per week)" = 2, "Trend Comparison" = 1))
```

## Pooled Regression vs. Gestational Age (GA)

For each cerebellar layer, the table below reports the **overall pooled OLS regression** of the layer measurement against **gestational age** (preterm + term combined; corresponds to the trend line shown in the left-hand panels of Figure 4 B/D/F/H). β is the slope per GA week, and 95% CI is the confidence interval on that slope.

```{r pooled-ga-regression}
pooled_ga_rows <- lapply(names(measurements), function(name) {
 m <- measurements[[name]]
 df <- data %>% filter(!is.na(.data[[m$col]]), !is.na(GA))
 fit <- lm(df[[m$col]] ~ df$GA)
 s <- summary(fit)$coefficients
 ci <- confint(fit)
 beta <- s["df$GA", "Estimate"]
 se <- s["df$GA", "Std. Error"]
 pval <- s["df$GA", "Pr(>|t|)"]
 tibble(
 Layer = name,
 `N` = nrow(df),
 `beta (per GA week)` = sprintf("%.3f", beta),
 `SE` = sprintf("%.3f", se),
 `95% CI` = sprintf("[%.3f, %.3f]", ci["df$GA", 1], ci["df$GA", 2]),
 `P` = format_pvalue(pval)
 )
})

bind_rows(pooled_ga_rows) %>%
 kable(caption = "Figure 4 (panels B/D/F/H): pooled OLS regression of each cerebellar layer on gestational age.",
 align = c("l","c","c","c","c","c")) %>%
 kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
```

The ML row corresponds to Figure 4F (Molecular Layer vs. Gestational Age).

---

# Conclusions

```{r conclusions, results='asis'}
# Generate conclusions dynamically from computed results
fmt_concl_p <- function(p) {
 if (is.na(p)) return("NA")
 if (p < 0.001) return("P < 0.001")
 sprintf("P = %.3f", p)
}

for (i in seq_len(nrow(results_table))) {
 layer <- results_table$Layer[i]
 mean_p <- fmt_concl_p(results_table$`Mean Diff p`[i])
 slope_p <- fmt_concl_p(results_table$`Slope Diff p`[i])

 sig_mean <- !is.na(results_table$`Mean Diff p`[i]) & results_table$`Mean Diff p`[i] < 0.05
 sig_slope <- !is.na(results_table$`Slope Diff p`[i]) & results_table$`Slope Diff p`[i] < 0.05

 mean_text <- if (sig_mean) "significantly different" else "not significantly different"
 slope_text <- if (sig_slope) {
 "a **significant difference in developmental trajectories**"
 } else {
 "similar developmental trajectories between groups"
 }

 cat(sprintf("%d. **%s**: Mean layer measurement is %s between Preterm and Term (%s), with %s (slope difference %s).\n\n",
 i, layer, mean_text, mean_p, slope_text, slope_p))
}
```

---

# Session Info

```{r session-info, echo=FALSE}
sessionInfo()
```
