---
title: "Figure 1C: Cerebellar Volume by Neurodevelopmental Delay Status"
subtitle: "Motor, Cognitive, Language — No Delay vs Delay (EP + VP), Wilcoxon test"
author: ""
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
 echo = TRUE, warning = FALSE, message = FALSE,
 fig.width = 3, fig.height = 4.5, dpi = 300
)
knitr::opts_knit$set(root.dir = normalizePath("."))

# --- Local dataset finder (RELATIVE paths only) ---------------------------
# Data lives in Figures_V2_Final/Supplementary_Datasets/ (git-ignored, kept off GitHub). This
# returns a RELATIVE path by walking up to find a "Supplementary_Datasets/" folder, so no
# machine-specific / user path is ever embedded in the code or rendered HTML.
find_dataset <- function(names, subdir = "Supplementary_Datasets", max_up = 6) {
 if (is.character(names) && length(names) == 1) names <- c(names)
 prefix <- "."
 for (i in 0:max_up) {
 for (nm in names) {
 rel <- file.path(prefix, subdir, nm)
 if (file.exists(rel)) return(rel)
 }
 prefix <- file.path(prefix, "..")
 }
 stop("Dataset not found under a 'Supplementary_Datasets/' folder: ", paste(names, collapse = " / "))
}
```

# Overview

Deck panel **Figure 1C**. Cerebellar volume (mm³) compared between infants **with vs without** delay,
for each neurodevelopmental domain (Motor, Cognitive, Language). Points are individual subjects coloured
by gestational-age group (EP / VP); LP excluded (low ND-outcome prevalence, EP+VP subset, GA < 32 wk).
Group comparison by two-sided **Wilcoxon rank-sum test**.

This is the boxplot analysis that was previously the *old* Figure S3
("Cerebellar volume vs neurodevelopmental delay"), promoted to main Figure 1C in the merged 4-figure deck.

```{r libs}
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(grid)
library(knitr)
library(kableExtra)
```

# Data (MRI volume × ND delay)

```{r load-merge}
# Colours
ga_colors <- c("Extremely Preterm" = "#030403", # ' updated palette (eyedropper)
 "Very Preterm" = "#DE1567",
 "Late Preterm" = "#7BACDC")
ga_labels_short <- c("Extremely Preterm" = "EP", "Very Preterm" = "VP", "Late Preterm" = "LP")

data_file <- find_dataset("Data_S1_InVivo_Cohort.xlsx")

# --- MRI (volumetric) ---
mri <- read_excel(data_file, sheet = "MRI") %>%
 filter(Included == "Yes") %>%
 rename(GA = GA_weeks, CerebVol = Cerebellar_Volume_mm3, GA_Group_raw = GA_Group) %>%
 mutate(
 Study_ID = as.character(Study_ID),
 GA_Category = factor(case_when(
 GA_Group_raw == "EP" ~ "Extremely Preterm",
 GA_Group_raw == "VP" ~ "Very Preterm",
 GA_Group_raw == "LP" ~ "Late Preterm"),
 levels = c("Extremely Preterm", "Very Preterm", "Late Preterm"))
 )

# --- ND (same cohort rule as Figure 1B) ---
nd <- read_excel(data_file, sheet = "Neurodevelopmental") %>%
 filter(Included == "Yes") %>%
 mutate(Study_ID = as.character(Study_ID))
# Exclude 6 patients with Delay="N" but all domain sub-scores NA (IDs 15,17,23,39,58,73)
nd <- nd %>% filter(!Study_ID %in% c("15","17","23","39","58","73"))

# --- Merge (subjects with BOTH volume and ND) ---
analysis_data <- mri %>%
 inner_join(nd %>% select(Study_ID, Any_Delay, Motor_Delay, Cognitive_Delay, Language_Delay),
 by = "Study_ID") %>%
 filter(Any_Delay %in% c("Y", "N"))

# Drop records with valid overall status but all-NA domains
invalid_domain <- analysis_data %>%
 filter(Any_Delay == "N", is.na(Motor_Delay) & is.na(Cognitive_Delay) & is.na(Language_Delay))
analysis_data <- analysis_data %>% anti_join(invalid_domain, by = "Study_ID")

# EP + VP subset (GA < 32 wk) — LP excluded
analysis_lt32 <- analysis_data %>% filter(GA < 32)

data.frame(
 Group = c("EP", "VP", "Total (EP+VP)"),
 N = c(sum(analysis_lt32$GA_Category == "Extremely Preterm"),
 sum(analysis_lt32$GA_Category == "Very Preterm"),
 nrow(analysis_lt32))
) %>% kable(caption = "Analytic cohort (volume × ND, GA < 32 wk)") %>%
 kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

# Figure 1C

```{r figure-1c, fig.width=3, fig.height=4.5}
# --- Long format: one row per subject × domain ---
long <- analysis_lt32 %>%
 select(Study_ID, CerebVol, GA_Category, Motor_Delay, Cognitive_Delay, Language_Delay) %>%
 pivot_longer(c(Motor_Delay, Cognitive_Delay, Language_Delay),
 names_to = "Domain", values_to = "Delay_Status") %>%
 filter(Delay_Status %in% c("N", "Y")) %>%
 mutate(
 Domain = dplyr::case_match(Domain, "Motor_Delay" ~ "Motor",
 "Cognitive_Delay" ~ "Cognitive", "Language_Delay" ~ "Language"),
 Domain = factor(Domain, levels = c("Motor", "Cognitive", "Language")),
 Delay_Status = factor(Delay_Status, levels = c("N", "Y"),
 labels = c("No\ndelay", "Delay"))
 )

# --- Wilcoxon per domain + bracket geometry ---
rng <- diff(range(long$CerebVol, na.rm = TRUE))
stat_df <- long %>%
 group_by(Domain) %>%
 summarise(p = wilcox.test(CerebVol ~ Delay_Status)$p.value,
 ymax = max(CerebVol, na.rm = TRUE), .groups = "drop") %>%
 mutate(
 bracket_y = ymax + rng * 0.06,
 tick = rng * 0.02,
 label_y = bracket_y + rng * 0.055,
 p_label = ifelse(p < 0.001, "italic(P) < 0.001",
 paste0("italic(P) == '", sprintf("%.3f", p), "'"))
 )

kable(stat_df %>% select(Domain, p), digits = 4,
 caption = "Wilcoxon rank-sum: volume, No delay vs Delay") %>%
 kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)

# Deck fill: No delay = white, Delay = warm tan/greige
fill_cols <- c("No\ndelay" = "white", "Delay" = "#DDD8D3")

# --- Single-panel layout: continuous x, ONE outer box (no facet dividers) ---
dom_lv <- c("Motor", "Cognitive", "Language")
dom_ctr <- c(1, 2.2, 3.4); names(dom_ctr) <- dom_lv
off <- 0.28
long$Domain <- factor(long$Domain, levels = dom_lv)
long$xc <- dom_ctr[as.character(long$Domain)]
long$xpos <- long$xc + ifelse(long$Delay_Status == "Delay", off, -off)

stat_df <- stat_df %>%
 mutate(Domain = factor(Domain, levels = dom_lv),
 xc = dom_ctr[as.character(Domain)], x_left = xc - off, x_right = xc + off)

y_lo <- min(long$CerebVol, na.rm = TRUE) - rng * 0.03
box_top <- 30000
ul_y <- box_top + rng * 0.075 # underline: clearly above the box
hdr_y <- box_top + rng * 0.105 # header text: just above its underline
hdr <- data.frame(xc = dom_ctr, lab = dom_lv)

fig_1c <- ggplot(long, aes(x = xpos, y = CerebVol)) +
 geom_boxplot(aes(group = xpos, fill = Delay_Status), width = 0.42, outlier.shape = NA,
 colour = "gray30", linewidth = 0.45) +
 geom_jitter(aes(colour = GA_Category), width = 0.10, alpha = 1, size = 1.1) +
 # significance brackets
 geom_segment(data = stat_df, inherit.aes = FALSE,
 aes(x = x_left, xend = x_right, y = bracket_y, yend = bracket_y), linewidth = 0.35, colour = "gray20") +
 geom_segment(data = stat_df, inherit.aes = FALSE,
 aes(x = x_left, xend = x_left, y = bracket_y - tick, yend = bracket_y), linewidth = 0.35, colour = "gray20") +
 geom_segment(data = stat_df, inherit.aes = FALSE,
 aes(x = x_right, xend = x_right, y = bracket_y - tick, yend = bracket_y), linewidth = 0.35, colour = "gray20") +
 geom_text(data = stat_df, inherit.aes = FALSE,
 aes(x = xc, y = label_y, label = p_label), parse = TRUE, size = 3.0, colour = "gray20") +
 # domain headers + underline (above the box)
 geom_text(data = hdr, inherit.aes = FALSE, aes(x = xc, y = hdr_y, label = lab), size = 3.4, vjust = 0) +
 geom_segment(data = hdr, inherit.aes = FALSE,
 aes(x = xc - 0.32, xend = xc + 0.32, y = ul_y, yend = ul_y), linewidth = 0.9) +
 scale_fill_manual(values = fill_cols, guide = "none") +
 scale_colour_manual(values = ga_colors, labels = ga_labels_short, name = NULL) +
 guides(colour = guide_legend(nrow = 1, override.aes = list(size = 2))) +
 scale_x_continuous(breaks = sort(unique(long$xpos)),
 labels = rep(c("No\ndelay", "Delay"), 3), expand = expansion(add = 0.4)) +
  scale_y_continuous(labels = label_comma(), breaks = seq(10000, 30000, 5000)) +
 coord_cartesian(ylim = c(y_lo, box_top), clip = "off") +
 labs(x = NULL, y = expression("Cerebellar volume (mm"^3*")")) +
 theme_classic(base_size = 10) +
 theme(
 axis.text.x = element_text(size = 10),
 axis.text.y = element_text(size = 10),
 axis.title.y = element_text(size = 10),
 axis.line = element_blank(),
 panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5),
 axis.ticks = element_line(linewidth = 0.3, colour = "gray30"),
 panel.grid = element_blank(),
 legend.position = c(0.015, 0.985),
 legend.justification = c(0, 1),
 legend.direction = "horizontal",
 legend.key.size = unit(0.28, "cm"),
 legend.text = element_text(size = 8),
 legend.background = element_rect(fill = "white", colour = "gray40", linewidth = 0.3),
 legend.margin = margin(2, 4, 2, 3),
 plot.margin = margin(t = 40, r = 6, b = 3, l = 3)
 )

fig_1c
ggsave("Figure_1C_Volume_by_Delay.pdf", fig_1c, width = 3.2, height = 4.5, dpi = 300)
message("Saved: Figure_1C_Volume_by_Delay.pdf")
```

# Session info

```{r session}
sessionInfo()
```
