---
title: "Figure S5 -- Twin vs singleton sensitivity"
subtitle: "Is multi-gestational status a confounder? (A) TEA cerebellar volume, (B) neurodevelopmental delay rates"
author: ""
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, dpi = 300)
knitr::opts_knit$set(root.dir = normalizePath("."))
library(readxl); library(dplyr); library(tidyr); library(ggplot2); library(knitr); library(kableExtra)

# --- Location-agnostic dataset finder (RELATIVE paths only) ---
find_dataset <- function(names, subdir = "Supplementary_Datasets", max_up = 6) {
 if (is.character(names) && length(names) == 1) names <- c(names)
 prefix <- "."
 for (i in 0:max_up) {
 for (nm in names) { rel <- file.path(prefix, subdir, nm); if (file.exists(rel)) return(rel) }
 prefix <- file.path(prefix, "..")
 }
 stop("Dataset not found under a 'Supplementary_Datasets/' folder: ", paste(names, collapse = " / "))
}
clin <- find_dataset("Data_S1_InVivo_Cohort.xlsx")

# Deck palette: Singleton = black, Twin = coral-red.
SINGLETON <- "#1A1A1A"; TWIN <- "#EE6677"
yn <- function(x) dplyr::case_match(toupper(trimws(x)), "Y" ~ 1L, "N" ~ 0L, .default = NA_integer_)
```

# Overview

Sensitivity analysis for whether **twin (multi-gestational) status** confounds the Figure 1 findings.
**A** TEA cerebellar volume, singleton vs twin (MRI cohort, Wilcoxon). **B** neurodevelopmental delay
rate (%) by twin status for Motor / Cognitive / Language (ND cohort, EP+VP; proportion ± 95% Wilson CI;
per-domain Fisher p, Holm-adjusted across the three domains). Panels exported as separate files.

# Panel A -- TEA cerebellar volume (singleton vs twin)

```{r panelA, fig.width=3.6, fig.height=4.4}
mri <- read_excel(clin, sheet = "MRI") %>% filter(Included == "Yes") %>%
 mutate(grp = factor(ifelse(Twin == "Y", "Twin", "Singleton"), levels = c("Singleton", "Twin")))
wv <- wilcox.test(Cerebellar_Volume_mm3 ~ grp, data = mri)
ymax <- max(mri$Cerebellar_Volume_mm3, na.rm = TRUE)

pA <- ggplot(mri, aes(grp, Cerebellar_Volume_mm3)) +
 geom_boxplot(width = 0.5, fill = NA, colour = "gray30", outlier.shape = NA, linewidth = 0.5) +
 geom_jitter(aes(colour = grp), width = 0.12, height = 0, size = 1.7, alpha = 0.6) +
 annotate("segment", x = 1, xend = 2, y = ymax * 1.06, yend = ymax * 1.06,
 colour = "gray20", linewidth = 0.4) +
 annotate("text", parse = TRUE, x = 1.5, y = ymax * 1.10,
 label = sprintf('italic(P) == "%.2f"', wv$p.value), size = 3.6, colour = "gray20") +
 scale_colour_manual(values = c(Singleton = SINGLETON, Twin = TWIN), guide = "none") +
 scale_y_continuous(breaks = seq(10000, 30000, 5000), limits = c(NA, ymax * 1.15)) +
 labs(x = NULL, y = expression(Cerebellar~volume~(mm^3))) +
 theme_classic(base_size = 12) +
 theme(axis.text = element_text(colour = "gray20"),
 axis.title.y = element_text(size = 11))
print(pA)
ggsave("Figure_S5A_Volume.pdf", pA, width = 3.6, height = 4.4, device = cairo_pdf, bg = "white")
```

# Panel B -- neurodevelopmental delay rate by twin status

```{r panelB, fig.width=3.5, fig.height=4.6}
nd <- read_excel(clin, sheet = "Neurodevelopmental") %>%
 filter(Included == "Yes", GA_Group %in% c("EP", "VP")) %>%
 mutate(grp = factor(ifelse(Twin == "Y", "Twin", "Singleton"), levels = c("Singleton", "Twin")))

wilson_ci <- function(k, n, z = 1.96) {
 p <- k / n; d <- 1 + z^2 / n
 lo <- (p + z^2/(2*n) - z*sqrt(p*(1-p)/n + z^2/(4*n^2))) / d
 hi <- (p + z^2/(2*n) + z*sqrt(p*(1-p)/n + z^2/(4*n^2))) / d
 c(lo = 100*lo, hi = 100*hi)
}
doms <- c(Motor = "Motor_Delay", Cognitive = "Cognitive_Delay", Language = "Language_Delay")

rate_df <- do.call(rbind, lapply(names(doms), function(dm) {
 col <- doms[[dm]]; x <- nd %>% filter(!is.na(.data[[col]]))
 x %>% group_by(grp) %>%
 summarise(k = sum(yn(.data[[col]])), n = n(), .groups = "drop") %>%
 rowwise() %>%
 mutate(rate = 100*k/n, lo = wilson_ci(k, n)[1], hi = wilson_ci(k, n)[2], domain = dm) %>%
 ungroup()
}))
rate_df$domain <- factor(rate_df$domain, levels = names(doms))

# per-domain Fisher (twin vs singleton), Holm-adjusted across the three domains
praw <- sapply(names(doms), function(dm) {
 col <- doms[[dm]]; x <- nd %>% filter(!is.na(.data[[col]]))
 fisher.test(table(x$grp, factor(yn(x[[col]]), levels = c(1, 0))))$p.value
})
padj <- p.adjust(praw, method = "holm")
yb <- 108
pdat <- data.frame(domain = factor(names(doms), levels = names(doms)), xx = seq_along(doms),
 p = padj)

dodge <- position_dodge(width = 0.55)
pB <- ggplot(rate_df, aes(domain, rate, colour = grp, group = grp)) +
 geom_errorbar(aes(ymin = lo, ymax = hi), position = dodge, width = 0.18, linewidth = 0.7) +
 geom_point(position = dodge, size = 3.0) +
 # per-domain P bracket (twin vs singleton, Holm-adjusted)
 geom_segment(data = pdat, inherit.aes = FALSE,
 aes(x = xx - 0.16, xend = xx + 0.16, y = yb, yend = yb),
 colour = "gray20", linewidth = 0.4) +
 geom_text(data = pdat, inherit.aes = FALSE, parse = TRUE,
 aes(x = xx, y = yb + 5, label = sprintf('italic(P) == "%.2f"', p)),
 size = 3.3, colour = "gray20") +
 scale_colour_manual(values = c(Singleton = SINGLETON, Twin = TWIN), name = NULL) +
 scale_x_discrete(expand = expansion(add = 0.85)) +
 scale_y_continuous(limits = c(0, 118), breaks = c(0, 25, 50, 75, 100)) +
 labs(x = NULL, y = "Delay rate (%)") +
 theme_classic(base_size = 12) +
 theme(legend.position = "top", legend.text = element_text(size = 10),
 axis.text = element_text(colour = "gray20"),
 axis.title.y = element_text(size = 11))
print(pB)
ggsave("Figure_S5B_DelayRates.pdf", pB, width = 3.5, height = 4.6, device = cairo_pdf, bg = "white")
```

# Numbers

```{r tbl, echo=FALSE}
bind_rows(
 tibble(Panel = "A volume (Wilcoxon)", Detail = sprintf("singleton n=%d, twin n=%d",
 sum(mri$grp=="Singleton"), sum(mri$grp=="Twin")),
 P = sprintf("%.2f", wv$p.value)),
 tibble(Panel = paste("B", names(doms), "delay (Fisher, Holm)"),
 Detail = sprintf("raw p=%.3f", praw), P = sprintf("%.2f", padj))
) %>% kable(caption = "Twin-vs-singleton comparisons") %>%
 kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
```

# Session Info

```{r session}
sessionInfo()
```
