---
title: "Figure 4C, D, G, H -- RNAScope ISH Validation"
subtitle: "H-Score Quantification of PLP1, KLF9, GRM4, TUBB4A Across Gestational Age"
author: ""
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
 echo = TRUE,
 message = FALSE,
 warning = FALSE,
 fig.width = 3.5,
 fig.height = 3,
 dpi = 300
)
knitr::opts_knit$set(root.dir = normalizePath("."))

# --- Local dataset finder (RELATIVE paths only; no user/machine path leaks) ---
find_dataset <- function(names, subdir = "Supplementary_Datasets", max_up = 6) {
 if (is.character(names) && length(names) == 1) names <- c(names)
 prefix <- "."
 for (i in 0:max_up) {
 for (nm in names) { rel <- file.path(prefix, subdir, nm); if (file.exists(rel)) return(rel) }
 prefix <- file.path(prefix, "..")
 }
 stop("Dataset not found under a 'Supplementary_Datasets/' folder: ", paste(names, collapse = " / "))
}
```

# Overview

Panels **C**, **D**, **G**, **H** present RNAScope *in situ* hybridization (ISH) quantification validating spatial transcriptomics findings. H-scores are plotted across gestational age for 4 genes in their respective cell populations.

| Panel | Gene | Cell Type | Colour | IF Channel |
|-------|------|-----------|--------|------------|
| **C** | PLP1 | Purkinje | `#4FD126` (green) | Opal570/Cy3 |
| **D** | KLF9 | Purkinje | `#CD2C22` (red) | Opal690/AF647 |
| **G** | GRM4 | Granule (IGL) | `#C82EDF` (magenta) | Opal620/AF594 |
| **H** | TUBB4A | Granule (IGL) | `#F0CA44` (yellow) | Opal520/AF488 |

# Libraries

```{r libraries}
library(readxl)
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)
```

---

# Source Data & Methods

## Data Provenance

RNAScope ISH was performed on **4 postmortem cerebellar specimens** using multiplex fluorescent panels. Images were acquired on a Leica microscope and quantified in QuPath.

**H-score** = (1 x %Low) + (2 x %Moderate) + (3 x %High), range 0--300, computed at the **ROI level**
(one score per Gene x Patient x ROI; ~5 ROIs per patient). Plots show individual ROI values with
mean +/- SEM. Granule-cell panels (G, H) use **IGL (Internal Granular Layer)** cells only.

**Input file**: `Supplementary_Datasets/Data_S5_InSitu_Validation.xlsx`, sheet `H_Score_ROI`.

## Data Loading

```{r load-data}
hscore_roi <- read_excel(find_dataset("Data_S5_InSitu_Validation.xlsx"), sheet = "H_Score_ROI")
```

## Source Data -- Summary

```{r source-summary}
hscore_roi %>%
 group_by(Gene, Cell_Type) %>%
 summarise(N_ROIs = n(), Mean_H = mean(H_score, na.rm = TRUE),
 SD_H = sd(H_score, na.rm = TRUE), .groups = "drop") %>%
 kable(digits = 1, caption = "Summary: ROI counts and H-scores per Gene and Cell Type") %>%
 kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

---

# Helpers

```{r helpers}
# Deck-matched styling: mean +/- SEM trend, plain upright title. No correlation coefficient is annotated
# (n = 4 subjects; ROIs are not independent replicates, so trends are shown descriptively),
# x-axis titled "Gestational age at birth (weeks)".
build_hscore_trend <- function(roi_data, gene_name, gene_col) {
 rd <- roi_data %>%
 group_by(GA) %>%
 summarise(
 mean_H = mean(H_score, na.rm = TRUE),
 se_H = sd(H_score, na.rm = TRUE) / sqrt(n()),
 .groups = "drop"
 ) %>%
 mutate(lo = pmax(mean_H - se_H, 0), hi = mean_H + se_H)


 ggplot() +
 geom_jitter(data = roi_data, aes(x = GA, y = H_score),
 color = gene_col, size = 1.5, alpha = 0.4, width = 0.3) +
 geom_errorbar(data = rd, aes(x = GA, ymin = lo, ymax = hi),
 color = gene_col, width = 1.5, linewidth = 0.6) +
 geom_line(data = rd, aes(x = GA, y = mean_H),
 color = gene_col, linewidth = 0.8) +
 geom_point(data = rd, aes(x = GA, y = mean_H),
 color = gene_col, size = 2.5) +
 scale_x_continuous(breaks = c(23, 28, 34.7, 40),
 labels = c("23", "28", "34.7", "40")) +
 labs(title = gene_name, x = "Gestational age at birth (weeks)", y = "H-score") +
 theme_classic(base_size = 11) +
 theme(
 plot.title = element_text(face = "plain", size = 11, hjust = 0.5),
 axis.title = element_text(size = 9),
 axis.text = element_text(size = 8)
 )
}

save_panel <- function(plt, stem) {
 ggsave(paste0(stem, ".pdf"), plt, width = 90, height = 80, units = "mm", dpi = 300)
}
```

---

# Figure 4C -- PLP1 (Purkinje)

```{r fig-4C}
plp1_purk <- hscore_roi %>% filter(Gene == "PLP1", Cell_Type == "Purkinje")
p4C <- build_hscore_trend(plp1_purk, "PLP1", gene_col = "#4FD126")
print(p4C); save_panel(p4C, "Figure_4C_PLP1_Hscore")
```

# Figure 4D -- KLF9 (Purkinje)

```{r fig-4D}
klf9_purk <- hscore_roi %>% filter(Gene == "KLF9", Cell_Type == "Purkinje")
p4D <- build_hscore_trend(klf9_purk, "KLF9", gene_col = "#CD2C22")
print(p4D); save_panel(p4D, "Figure_4D_KLF9_Hscore")
```

# Figure 4G -- GRM4 (Granule, IGL)

```{r fig-4G}
grm4_gc <- hscore_roi %>% filter(Gene == "GRM4", Cell_Type == "IGL")
p4G <- build_hscore_trend(grm4_gc, "GRM4", gene_col = "#C82EDF")
print(p4G); save_panel(p4G, "Figure_4G_GRM4_Hscore")
```

# Figure 4H -- TUBB4A (Granule, IGL)

```{r fig-4H}
tubb4a_gc <- hscore_roi %>% filter(Gene == "TUBB4A", Cell_Type == "IGL")
p4H <- build_hscore_trend(tubb4a_gc, "TUBB4A", gene_col = "#F0CA44")
print(p4H); save_panel(p4H, "Figure_4H_TUBB4A_Hscore")
```

---

# Session Info

```{r session-info}
sessionInfo()
```
