---
title: "Figure 1B: Neurodevelopmental Delay Prevalence (EP vs VP)"
subtitle: "Motor, Cognitive, and Language Delay — Fisher's Exact Test"
author: ""
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
 echo = TRUE, warning = FALSE, message = FALSE,
 fig.width = 3.5, fig.height = 5, dpi = 300
)
knitr::opts_knit$set(root.dir = normalizePath("."))

# --- Local dataset finder (RELATIVE paths only) ---------------------------
# Data lives in Figures_V2_Final/Supplementary_Datasets/ (git-ignored, kept off GitHub). This
# returns a RELATIVE path by walking up to find a "Supplementary_Datasets/" folder, so no
# machine-specific / user path is ever embedded in the code or rendered HTML.
find_dataset <- function(names, subdir = "Supplementary_Datasets", max_up = 6) {
 if (is.character(names) && length(names) == 1) names <- c(names)
 prefix <- "."
 for (i in 0:max_up) {
 for (nm in names) {
 rel <- file.path(prefix, subdir, nm)
 if (file.exists(rel)) return(rel)
 }
 prefix <- file.path(prefix, "..")
 }
 stop("Dataset not found under a 'Supplementary_Datasets/' folder: ", paste(names, collapse = " / "))
}
```

# Data Loading and Exclusion Flow

## Cohort Derivation

The neurodevelopmental cohort with valid assessments (N = 169, including LP) was derived from the full enrolled sample. This panel analyzes the extremely-preterm and very-preterm subset (N = 145; LP excluded). No MRI data is required for this panel.

```{r exclusion_flow}
library(readxl)
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)

# Exclusion flow table
excl_flow <- data.frame(
 Step = c(
 "MRI scans after quality review",
 "Scan-timing exclusions",
 "",
 "After scan-timing exclusions",
 "Late-preterm ND-only exclusion",
 "After LP ND-only exclusion",
 "No valid ND assessment in 9-36 month window",
 "Invalid domain records excluded",
 "**Final ND cohort (incl. LP)**"
 ),
 Detail = c(
 "",
 "ID 469 (GA=33.8, LP): Scanned ~2 days after birth (PMA 34.1 wk), not TEA",
 "ID 57 (GA=28.0, VP): Scanned 20 wk post-birth (PMA 48 wk), outside TEA window",
 "",
 "IDs 140, 478 (LP; valid ND but no MRI or ultrasound; outside all analyzed modalities)",
 "",
 "ID 465 (GA=30.5, VP): Only assessment in window has missing (NaN) Delay value",
 "6 subjects (IDs 15, 17, 23, 39, 58, 73) with Delay=N but all domain sub-scores NA -- excluded",
 ""
 ),
 N = c("N = 180", "-1", "-1", "N = 178", "-2", "N = 176", "-1", "-6", "**N = 169**"),
 stringsAsFactors = FALSE
)

kable(excl_flow,
 caption = "Exclusion Flow: Neurodevelopmental Cohort",
 col.names = c("Step", "Detail", "N")) %>%
 kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
 row_spec(9, bold = TRUE, background = "#f0f7fb")
```

**ND Assessment Window:** 9--36 months corrected age. When multiple assessments exist, the one closest to the 22.5-month midpoint with a valid Delay classification (Y/N) is selected.

**GA Group Definitions:**

| Group | Label | GA Range |
|-------|-------|----------|
| EP | Extremely Preterm | < 28 weeks |
| VP | Very Preterm | 28 to < 32 weeks |
| LP | Late Preterm | >= 32 weeks |

**Data Source:** `Supplementary_Datasets/Data_S1_InVivo_Cohort.xlsx`, sheet = "Neurodevelopmental". The sheet contains all 178 subjects (post scan-timing exclusions); excluded subjects are marked via the `Included` and `Exclusion_Reason` columns. The code below filters to `Included == "Yes"` to obtain the analytic sample.

## Load Data

```{r load_data}
# ============================================================
# LOAD FROM FINAL ANALYSIS DATASET — Neurodevelopmental sheet
# Sheet contains all 178 subjects; filter to Included == "Yes"
# for the with-LP ND cohort (N = 169); Fig 1B uses the EP+VP subset (N = 145)
# ============================================================
data_file <- find_dataset("Data_S1_InVivo_Cohort.xlsx")
nd_all <- read_excel(data_file, sheet = "Neurodevelopmental")

# Show what was excluded
excluded <- nd_all %>% filter(Included == "No")
if (nrow(excluded) > 0) {
 excluded %>%
 select(Study_ID, GA_weeks, GA_Group, Exclusion_Reason) %>%
 kable(caption = "Excluded from Neurodevelopmental Analysis",
 col.names = c("Study ID", "GA (wk)", "Group", "Reason")) %>%
 kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
}

# Filter to included subjects only
nd_data <- nd_all %>% filter(Included == "Yes")

# Exclude 6 patients with Delay="N" but ALL domain sub-scores NA
# (IDs 15, 17, 23, 39, 58, 73) — consistent with the original analysis pipeline
domain_invalid_ids <- c(15, 17, 23, 39, 58, 73)
nd_data <- nd_data %>% filter(!Study_ID %in% domain_invalid_ids)

# Create binary columns
nd_data <- nd_data %>%
 mutate(
 GA_Category = factor(
 case_when(
 GA_weeks < 28 ~ "Extremely Preterm",
 GA_weeks < 32 ~ "Very Preterm",
 TRUE ~ "Late Preterm"
 ),
 levels = c("Extremely Preterm", "Very Preterm", "Late Preterm")
 ),
 Motor_Delay_Binary = ifelse(Motor_Delay == "Y", 1L, ifelse(Motor_Delay == "N", 0L, NA_integer_)),
 Cognitive_Binary = ifelse(Cognitive_Delay == "Y", 1L, ifelse(Cognitive_Delay == "N", 0L, NA_integer_)),
 Language_Delay_Binary = ifelse(Language_Delay == "Y", 1L, ifelse(Language_Delay == "N", 0L, NA_integer_))
 )

# Color scheme (' updated palette 2026-08 — eyedropper from PowerPoint)
ga_colors <- c(
 "Extremely Preterm" = "#030403", # EP near-black
 "Very Preterm" = "#DE1567", # VP crimson/magenta
 "Late Preterm" = "#7BACDC" # LP light blue
)

ga_fill <- ga_colors # solid fills (deck: filled bars)

# Cohort summary
data.frame(
 Metric = c("Total ND cohort", "EP (<28 wk)", "VP (28-<32 wk)", "LP (>=32 wk)"),
 N = c(nrow(nd_data),
 sum(nd_data$GA_Category == "Extremely Preterm"),
 sum(nd_data$GA_Category == "Very Preterm"),
 sum(nd_data$GA_Category == "Late Preterm"))
) %>%
 kable(caption = "Neurodevelopmental Cohort Summary",
 col.names = c("Group", "N")) %>%
 kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# Figure 1B: ND Delay Prevalence (EP vs VP)

```{r figure_1b, fig.width=3.5, fig.height=5}
# ============================================================
# FIGURE 1B: Grouped bar chart — Motor, Cognitive, Language
# EP vs VP only (LP excluded from this panel)
# Fisher's exact test with significance brackets
# Matching the original pipeline methods exactly
# ============================================================

# --- Helper: format p-value ---
format_pval <- function(p) {
 if (is.na(p)) return("NA")
 if (p < 0.001) return("P < 0.001")
 if (p < 0.01) return(paste0("P = ", formatC(p, format = "f", digits = 3)))
 if (p < 0.05) return(paste0("P = ", formatC(p, format = "f", digits = 3)))
 return(paste0("P = ", formatC(p, format = "f", digits = 2)))
}

# --- Subset to EP and VP only ---
ep_vp_data <- nd_data %>%
 filter(GA_Category %in% c("Extremely Preterm", "Very Preterm")) %>%
 mutate(GA_Category = droplevels(GA_Category))

data.frame(
 Group = c("EP", "VP", "Total (EP + VP)"),
 N = c(sum(ep_vp_data$GA_Category == "Extremely Preterm"),
 sum(ep_vp_data$GA_Category == "Very Preterm"),
 nrow(ep_vp_data))
) %>%
 kable(caption = "EP + VP Subset for Figure 1B", col.names = c("Group", "N")) %>%
 kable_styling(bootstrap_options = c("condensed"), full_width = FALSE)

# --- Delay domains ---
delay_info <- list(
 list(col = "Motor_Delay_Binary", label = "Motor"),
 list(col = "Cognitive_Binary", label = "Cognitive"),
 list(col = "Language_Delay_Binary", label = "Language")
)

# --- Build prevalence data + Fisher's tests ---
all_prev <- data.frame()
fisher_results <- list()

for (info in delay_info) {
 col <- info$col
 label <- info$label

 prev_data <- ep_vp_data %>%
 filter(!is.na(.data[[col]])) %>%
 group_by(GA_Category) %>%
 summarise(
 N = n(),
 N_Delay = sum(.data[[col]], na.rm = TRUE),
 Rate = mean(.data[[col]], na.rm = TRUE) * 100,
 .groups = "drop"
 ) %>%
 mutate(
 SE = sqrt(Rate/100 * (1 - Rate/100) / N) * 100,
 CI_lower = pmax(Rate - 1.96 * SE, 0),
 CI_upper = pmin(Rate + 1.96 * SE, 100),
 Domain = label
 )

 all_prev <- bind_rows(all_prev, prev_data)

 # Fisher's exact test
 ep_vp_valid <- ep_vp_data %>%
 filter(!is.na(.data[[col]])) %>%
 mutate(GA_Category = droplevels(GA_Category))
 tbl <- table(ep_vp_valid$GA_Category, ep_vp_valid[[col]])
 fish_test <- tryCatch(fisher.test(tbl), error = function(e) list(p.value = NA))
 fisher_results[[label]] <- fish_test$p.value
}

# Set factor levels for domain ordering
all_prev$Domain <- factor(all_prev$Domain, levels = c("Motor", "Cognitive", "Language"))

# GA group labels for legend
ga_bar_labels <- c("Extremely Preterm" = "EP",
 "Very Preterm" = "VP")
all_prev$GA_Label <- ga_bar_labels[as.character(all_prev$GA_Category)]
all_prev$GA_Label <- factor(all_prev$GA_Label, levels = ga_bar_labels)

# --- Rename Domain levels to include "delay" (matching reference) ---
all_prev$Domain <- factor(
 paste0(all_prev$Domain, "\ndelay"),
 levels = c("Motor\ndelay", "Cognitive\ndelay", "Language\ndelay")
)

# --- Build Fisher p-value label per domain ---
fisher_labels <- data.frame(
 Domain = factor(
 paste0(names(fisher_results), "\ndelay"),
 levels = levels(all_prev$Domain)
 ),
 p_val = sapply(fisher_results, identity),
 p_label = sapply(fisher_results, function(p) {
 if (is.na(p)) return("")
 if (p < 0.001) return("italic(P) < 0.001")
 # quote the value so plotmath keeps trailing zeros (e.g. 0.030, not 0.03)
 paste0("italic(P) == '", sprintf("%.3f", p), "'")
 }),
 stringsAsFactors = FALSE
)

# --- Single-panel grouped bar chart: ONE outer box (no internal dividers) ---
# Continuous x positions: 3 domains spaced apart, EP/VP dodged within each.
dom_lv <- c("Motor\ndelay", "Cognitive\ndelay", "Language\ndelay")
dom_ctr <- c(1, 2.2, 3.4); names(dom_ctr) <- dom_lv # domain centres (gap between domains)
off <- 0.28 # EP/VP offset within a domain
all_prev$Domain <- factor(all_prev$Domain, levels = dom_lv)
all_prev$xc <- dom_ctr[as.character(all_prev$Domain)]
all_prev$xpos <- all_prev$xc + ifelse(all_prev$GA_Label == "EP", -off, off)
xdf <- all_prev[order(all_prev$xpos), ]
y_top <- 95

# P brackets per domain (over the EP/VP pair)
domain_max_ci <- all_prev %>% group_by(Domain) %>%
 summarise(max_ci = max(CI_upper, na.rm = TRUE), xc = first(xc), .groups = "drop")
bracket_data <- fisher_labels %>%
 left_join(domain_max_ci, by = "Domain") %>%
 filter(!is.na(p_val) & p_val < 0.10) %>%
 mutate(x_left = xc - off, x_right = xc + off, bracket_y = max_ci + 6, label_y = bracket_y + 3)

# domain headers (above the box) + underline
hdr <- data.frame(xc = dom_ctr, lab = dom_lv)

fig_1c <- ggplot(all_prev, aes(x = xpos, y = Rate, fill = GA_Label)) +
 geom_col(width = 0.5, colour = NA) +
 geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper),
 width = 0.14, linewidth = 0.5, colour = "black") +
 # significance brackets
 geom_segment(data = bracket_data, inherit.aes = FALSE,
 aes(x = x_left, xend = x_right, y = bracket_y, yend = bracket_y), linewidth = 0.35) +
 geom_segment(data = bracket_data, inherit.aes = FALSE,
 aes(x = x_left, xend = x_left, y = bracket_y - 1.5, yend = bracket_y), linewidth = 0.35) +
 geom_segment(data = bracket_data, inherit.aes = FALSE,
 aes(x = x_right, xend = x_right, y = bracket_y - 1.5, yend = bracket_y), linewidth = 0.35) +
 geom_text(data = bracket_data, inherit.aes = FALSE,
 aes(x = xc, y = label_y, label = p_label), parse = TRUE, size = 3.0) +
 # domain headers + underline (above the box; clip off) — underline sits just under the text
 geom_text(data = hdr, inherit.aes = FALSE, aes(x = xc, y = y_top + 15, label = lab),
 size = 3.4, lineheight = 0.9, vjust = 1) +
 geom_segment(data = hdr, inherit.aes = FALSE,
 aes(x = xc - 0.5, xend = xc + 0.5, y = y_top + 6, yend = y_top + 6), linewidth = 0.9) +
 scale_fill_manual(values = c("EP" = ga_colors[["Extremely Preterm"]],
 "VP" = ga_colors[["Very Preterm"]]), guide = "none") +
 scale_x_continuous(breaks = xdf$xpos, labels = as.character(xdf$GA_Label),
 expand = expansion(add = 0.45)) +
 scale_y_continuous(breaks = seq(0, 80, 20), expand = expansion(mult = c(0, 0.05))) +
 coord_cartesian(ylim = c(0, y_top), clip = "off") +
 labs(x = NULL, y = "Prevalence (%)") +
 theme_classic(base_size = 10) +
 theme(
 axis.text.x = element_text(size = 10),
 axis.text.y = element_text(size = 10),
 axis.title.y = element_text(size = 10),
 axis.line = element_blank(),
 panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5),
 axis.ticks = element_line(linewidth = 0.3, color = "gray30"),
 plot.margin = margin(t = 42, r = 6, b = 2, l = 2)
 )

fig_1c
ggsave("Figure_1B_ND_Prevalence.pdf", fig_1c, width = 2.9, height = 4.5, dpi = 300)
message("Saved: Figure_1B_ND_Prevalence.pdf")
```

## Prevalence Summary Table

```{r prevalence_table}
prev_table <- all_prev %>%
 mutate(
 Rate_CI = sprintf("%.1f%% (%.1f-%.1f)", Rate, CI_lower, CI_upper),
 Fraction = paste0(N_Delay, "/", N)
 ) %>%
 select(Domain, GA_Label, Fraction, Rate_CI)

kable(prev_table,
 caption = "Figure 1B: ND Delay Prevalence — EP vs VP",
 col.names = c("Domain", "Group", "n/N", "Rate (95% CI)")) %>%
 kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Fisher's Exact Test Results

```{r fisher_table}
fish_table <- data.frame(
 Domain = names(fisher_results),
 P_Value = sapply(fisher_results, function(p) {
 if (is.na(p)) "NA"
 else if (p < 0.001) "< 0.001"
 else sprintf("%.4f", p)
 }),
 Significant = sapply(fisher_results, function(p) ifelse(!is.na(p) & p < 0.05, "Yes", "No")),
 row.names = NULL
)

kable(fish_table,
 caption = "Fisher's Exact Test: EP vs VP by Domain",
 col.names = c("Domain", "p-value", "Significant (p < 0.05)")) %>%
 kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# Session Info

```{r session_info}
sessionInfo()
```
