---
title: "Figure 1D: Transcerebellar Diameter Z-score Trajectory vs Postmenstrual Age"
subtitle: "Serial cranial ultrasound TCD referenced to INTERGROWTH-21st (ig_fet / tcdfga); EP / VP / LP"
author: ""
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
 fig.width = 2.2, fig.height = 3.0, dpi = 300)
knitr::opts_knit$set(root.dir = normalizePath("."))

# --- Local dataset finder (RELATIVE paths only) ---------------------------
# Data lives in Figures_V2_Final/Supplementary_Datasets/ (git-ignored). Returns a RELATIVE path
# so no machine/user path is embedded in code or rendered HTML.
find_dataset <- function(names, subdir = "Supplementary_Datasets", max_up = 6) {
 if (is.character(names) && length(names) == 1) names <- c(names)
 prefix <- "."
 for (i in 0:max_up) {
 for (nm in names) { rel <- file.path(prefix, subdir, nm); if (file.exists(rel)) return(rel) }
 prefix <- file.path(prefix, "..")
 }
 stop("Dataset not found under a 'Supplementary_Datasets/' folder: ", paste(names, collapse = " / "))
}
```

# Overview

Deck panel **Figure 1D** (row "Serial cranial ultrasound"). The reviewer asked us to compare our
preterm infants directly against a **published in-utero fetal normative standard**, like-for-like, so the
deviation from the expected fetal trajectory is shown explicitly.

We use the **transcerebellar diameter (TCD)**, measured in the coronal plane on serial cranial ultrasound
(`Coronal_Transverse` in the dataset, in mm), and reference each measurement to the **INTERGROWTH-21st**
international fetal standard for cerebellar growth (Rodriguez-Sibaja et al., *Ultrasound Obstet Gynecol*
2021;57:614-623; PMID 32196791), via the `gigs` package (`family = "ig_fet"`, `acronym = "tcdfga"`).

Because fetal gestational age and postnatal postmenstrual age (PMA) are the **same clock**, each infant scan
at *N* weeks PMA is compared to the fetal norm at GA = *N* weeks. TCD is expressed as a **z-score** and plotted
against PMA. The INTERGROWTH standard extends to ~40 weeks, so scans beyond **280 days (40 wk) PMA** have no
reference and are dropped (per agreement with C. Cherry). This analysis reproduces `caller.R` (C. Cherry).

```{r libs}
library(gigs)
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(mgcv)
library(lme4)
library(lmerTest)
library(knitr)
```

# Data + z-score conversion

```{r load}
days_per_month <- 30.44
grp_order <- c("EP", "VP", "LP")
group_cols <- c(EP = "#030403", VP = "#DE1567", LP = "#7BACDC") # ' EP/VP/LP palette

data_file <- find_dataset(c("Dataset_Clinical_MRI_US_ND.xlsx",
 "Dataset_S1_Clinical_MRI_US_ND.xlsx"))

exp <- read_excel(data_file, sheet = "Ultrasound") %>%
 # TCD (transcerebellar diameter, mm) = coronal transverse cerebellar measurement
 mutate(
 TCD = as.numeric(Coronal_Transverse),
 PMA_weeks = as.numeric(PMA_weeks),
 PMA_days = PMA_weeks * 7,
 PMA_months = PMA_days / days_per_month,
 Study_ID = as.character(Study_ID)
 ) %>%
 filter(PMA_days <= 280 & !is.na(TCD)) %>%
 mutate(GA_Group = factor(GA_Group, levels = grp_order))

# Convert each measured TCD to a z-score vs the INTERGROWTH-21st fetal standard
exp$Z.Score <- value2zscore(
 exp$TCD,
 exp$PMA_days,
 family = "ig_fet",
 acronym = "tcdfga"
)

kable(exp %>% count(GA_Group, name = "N_measurements"),
 caption = "TCD measurements per GA group (PMA <= 280 days)")
```

# Statistics

```{r stats-zpermonth}
# 1. Change in z-score per month (slope; negative = falling further below the norm).
# The slope p-value is scale-invariant, so it is identical whether PMA is in
# weeks or months; annotations report this p-value per group.
growth_rate <- exp %>%
 group_by(GA_Group) %>%
 summarise(
 Z_per_Month = coef(lm(Z.Score ~ PMA_months))[2],
 p = summary(lm(Z.Score ~ PMA_months))$coef[2, 4],
 .groups = "drop"
 )
kable(growth_rate, digits = c(0, 3, 6),
 caption = "Change in TCD z-score per month, by GA group")
```

```{r stats-below5}
# 2. Proportion of measurements below the reference 5th percentile (z < -1.645).
below5 <- exp %>%
 group_by(GA_Group) %>%
 summarise(
 N = n(),
 Below5 = sum(Z.Score < -1.645),
 Percent = 100 * mean(Z.Score < -1.645),
 .groups = "drop"
 )
kable(below5, digits = c(0, 0, 0, 1),
 caption = "Percent of measurements below the reference 5th percentile")
```

```{r stats-maxdeficit}
# 3. Maximum smoothed z-score deficit (GAM peak deficit), for reference.
pred <- expand_grid(
 PMA_months = seq(98, 280) / days_per_month,
 GA_Group = factor(grp_order, levels = grp_order)
)
gam_fit <- gam(Z.Score ~ GA_Group + s(PMA_months, by = GA_Group),
 data = exp, method = "REML")
pred$Z <- predict(gam_fit, newdata = pred)
max_deficit <- pred %>%
 group_by(GA_Group) %>%
 slice_min(Z, n = 1, with_ties = FALSE) %>%
 transmute(GA_Group,
 Peak_Deficit_Month = PMA_months,
 Peak_Deficit_Z = Z)
kable(max_deficit, digits = c(0, 2, 2),
 caption = "Maximum smoothed z-score deficit (GAM)")
```

```{r stats-lme}
# 4. Overall trajectory difference across groups: linear mixed-effects model with
# fixed effects for PMA, group and their interaction, random intercept per infant,
# Type III tests with Satterthwaite degrees of freedom.
fit <- lmer(TCD ~ PMA_months * GA_Group + (1 | Study_ID), data = exp)
anova(fit)
```

# Figure 1D

```{r figure, fig.width=2.2, fig.height=3.0}
# Annotation strings built from the computed statistics ---------------------
fmt_p <- function(p) ifelse(p < 0.001, "italic(P) < 0.001",
 paste0("italic(P) == '", sprintf("%.3f", p), "'"))
lab <- below5 %>%
 left_join(growth_rate, by = "GA_Group") %>%
 mutate(GA_Group = factor(GA_Group, levels = grp_order)) %>%
 arrange(GA_Group) %>%
 mutate(
 txt = paste0(as.character(GA_Group), "*' – '*", fmt_p(p),
 "*'; '*", sprintf("%.0f", Percent), "*'% <5th'"),
    y   = c(-3.50, -3.85, -4.20)[seq_len(n())]
 )

x_lo <- 22; x_hi <- 40
ref_x <- x_hi + 0.15 # ref-line labels sit just past the right edge

# In-plot legend (top-left), coloured line keys + labels (matches deck panel D)
leg <- data.frame(
 GA_Group = factor(grp_order, levels = grp_order),
 y = c(1.55, 1.00, 0.45)
)

p1d <- ggplot(exp, aes(PMA_weeks, Z.Score, color = GA_Group, fill = GA_Group)) +
 # Reference lines: fetal median and +/- 2 SD (97.7th / 2.3rd centiles)
 geom_hline(yintercept = 2, linetype = "dashed", colour = "grey55", linewidth = 0.3) +
 geom_hline(yintercept = 0, linetype = "dashed", colour = "grey35", linewidth = 0.3) +
 geom_hline(yintercept = -2, linetype = "dotted", colour = "grey55", linewidth = 0.3) +
 # Per-group linear fits vs PMA with 95% CI
 geom_smooth(method = "lm", se = TRUE, linewidth = 0.9, alpha = 0.18) +
 scale_color_manual(values = group_cols) +
 scale_fill_manual(values = group_cols) +
 # Reference-line labels (right side)
 annotate("text", x = ref_x, y = 2, label = "97.7th", hjust = 0, size = 2.5,
 fontface = "italic", colour = "grey40") +
 annotate("text", x = ref_x, y = 0, label = "median", hjust = 0, size = 2.5,
 fontface = "italic", colour = "grey30") +
 annotate("text", x = ref_x, y = -2, label = "2.3rd", hjust = 0, size = 2.5,
 fontface = "italic", colour = "grey40") +
 # Per-group statistic annotations (bottom-left), coloured by group
 geom_text(data = lab, aes(x = x_lo + 0.4, y = y, label = txt, colour = GA_Group),
 parse = TRUE, hjust = 0, size = 2.5, inherit.aes = FALSE,
 show.legend = FALSE) +
 # In-plot legend, top-left: coloured line keys + EP/VP/LP labels
 geom_segment(data = leg, aes(x = x_lo + 0.5, xend = x_lo + 1.9, y = y, yend = y,
 colour = GA_Group),
 linewidth = 0.9, inherit.aes = FALSE, show.legend = FALSE) +
 geom_text(data = leg, aes(x = x_lo + 2.2, y = y, label = as.character(GA_Group)),
 hjust = 0, size = 2.6, colour = "black", inherit.aes = FALSE) +
 scale_x_continuous(breaks = c(22, 28, 33, 39), limits = c(x_lo, x_hi + 3)) +
 scale_y_continuous(breaks = c(2, 0, -2, -4)) +
 coord_cartesian(xlim = c(x_lo, x_hi), ylim = c(-4.4, 2.3),
 expand = FALSE, clip = "off") +
 labs(x = "Post-menstrual age (weeks)",
 y = "Transcerebellar diameter\n(Z-score from reference median)") +
 theme_classic(base_size = 10) +
 theme(
 legend.position = "none",
 axis.text.x = element_text(size = 9),
 axis.text.y = element_text(size = 9),
 axis.title.x = element_text(size = 9),
 axis.title.y = element_text(size = 9),
 axis.line = element_blank(),
 panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5),
 axis.ticks = element_line(linewidth = 0.3, colour = "black"),
 panel.grid = element_blank(),
 plot.margin = margin(6, 26, 4, 4)
 )

p1d
ggsave("Figure_1D_TCD_Zscore_Trajectory.pdf", p1d, width = 2.2, height = 3.0, dpi = 300)
```

# Session info

```{r session}
sessionInfo()
```
